### 2019-04-10 mpc19/paper.tex
first commit

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\usepackage{proof}
\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathtools}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Pio}{$\Pi^{\mathsf{o}}$} 
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\comp}{\circ}

% Delay macros
\newcommand{\Delay}{\mathsf{Delay}\,}
\newcommand{\now}{\mathsf{now}}
\newcommand{\later}{\mathsf{later}}
\newcommand{\never}{\mathsf{never}}
\newcommand{\dn}{\downarrow}
\newcommand{\bind}{\mathsf{bind}}
\newcommand{\str}{\mathsf{str}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\Dapprox}{\mathbb{D}_{\approx}}

\begin{document}
%
\title{Type-theoretic Semantics of \Pio\ via the Delay Monad \thanks{Supported by organization x.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Robin Kaarsgaard \inst{1}\orcidID{0000-0002-7672-799X} \and
Niccol\`o Veltri\inst{2}\orcidID{0000-0002-7230-3436}}
%
\authorrunning{Kaarsgaard and Veltri}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{DIKU, Department of Computer Science, University of
    Copenhagen\\ \email{robin@di.ku.dk} \and
Department of Computer Science, IT University of
    Copenhagen\\ \email{nive@itu.dk}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}

\section{Introduction}\label{sec:intro}

\section{Syntax of \Pio}\label{sec:syntax}

\section{Delay Monad}\label{sec:delay}

The coinductive delay datatype was first introduce by Capretta for
representing general recursive functions in Martin-L\"of type theory
\cite{Capretta05}.  Given a type $A$, elements of $\Delay A$ are
possibly non-terminating ``computations'' returning a value of $A$
whenever they terminate. Formally, $\Delay A$ is defined as a
coinductive type with the following introduction rules:
\[
\infer={\now\,a : \Delay A}{x : A}
\quad
\infer={\later\,x : \Delay A}{x : \Delay A}
\]
(Here and in the following, single rule lines refer to an inductive
definition, double rule lines to a coinductive definition.) The
constructor $\now$ embeds $A$ into $\Delay A$, so $\now\,a$ represents
the terminating computation returning the value $a$. The constructor
$\later$ adds an additional unit of time delay to a computation. The
non-terminating computation $\never$ is corecursively defined as
$\never = \later \,\never$.

The delay datatype is a monad. The unit is the constructor $\now$,
while the Kleisli extension $\bind$ is corecursively defined as follows:
\begin{align*}
& \bind : (A \to \Delay B) \to \Delay A \to \Delay B \\
& \bind \,f \, (\now\,a) = f\,a \\
& \bind\,f\,(\later\,x) = \later\,(\bind\,f\,x)
\end{align*}
The delay monad, like any other monad on $\Set$, has a unique strength
operation which we denote by $\str : A \times \Delay B \to \Delay (A
\times B)$. Moreover, the delay datatype is a commutative monad.

The Kleisli category of the delay monad, that we call $\D$, has types
as objects and functions $f : A \to \Delay B$ as morphisms between $A$
and $B$. In $\D$, the identity map on an object $A$ is the constructor
$\now$, while the composition of morphisms $f : A \to \Delay B$ and $g
: B \to \Delay C$ is given by $f \diamond g = \bind\,f \comp g$.

The delay datatype allows us to program with partial functions, but
the introduced notion of partiality is intensional, in the sense that
computations terminating with the same value in a different number of
steps are considered different. To obtain an extensional notion of
partiality, which in particular allows the specification of a
well-behaved trace operator, we need to quotient the delay datatype by
weak bisimilarity.

Weak bisimilarity is defined in terms of convergence. A computation
$x : \Delay A$ converges to $a : A$ if it terminates in a finite
number of steps returning the value $a$. When this happens, we write
$x \dn a$. The relation $\dn$ is inductively defined by the rules:
\[
\infer{\now\,a \dn a}{}
\quad
\infer{\later\,x \dn a}{x \dn a}
\]
Two computations in $\Delay A$ are weakly bisimilar if they differ by
a finite number of applications of the constructor $\later$. This
informal statement can be formalized in several different but
logically equivalent ways \cite{ChapmanUV19}. For this work we chose the
following definition:
\[
x \approx y = (a : A) \to (x \dn a \to y \dn a) \times (y \dn a \to x \dn a) 
\]
The definition says that $x$ and $y$ are weakly bisimilar if, whenever
$x$ terminates, then $y$ also terminates with the same value, and vice
versa. Weak bisimilarity is an equivalence relation and it is a
congrence with respect to the $\later$ operation.  We call $\Dapprox$
the Kleisli category of $\Delay$ with homsets quotiented by pointwise
weak bisimilarity. This means that in $\Dapprox$ two morphisms $f$ and
$g$ are considered equal whenever $f \, a \approx g \, a$, for all
inputs $a$. When this is the case, we also write $f \approx g$.  In
this work we do not formally construct the quotients
$(A \to \Delay B)/{\approx}$, since we work in a type theory without
quotient types. Instead we require that every construction we perform
on morphisms in $\Dapprox$ respects weak bisimilarity. Technically, we
construct $\Dapprox$ as a category enriched in setoids CITE, where in
particular all the reasoning on morphisms is performed up-to weak
bisimilarity.

In previous work \cite{ChapmanUV19} we showed that, when working in a
type theory with quotient types a la Hofmann CITE, quotienting the
delay monad by weak bisimilarity has problematic consequences. In
fact, it does not seem possible to recover the monad structure on the
quotiented datatype $\mathsf{DelayQuot}\,A = \Delay A/{\approx}$
without assuming additional controversial principles
such as the axiom of countable choice. In particolar, countable choice
would be necessary for modelling the trace operator of \Pio. In
loc.cit. we showed that a possible way to avoid these issues is
forming the quotients of homsets, as in $(A \to \Delay B)/{\approx}$, instead of
quotienting the delay datatype directly. We would still not be able to form

\section{Elgot Iteration and Trace}\label{sec:elgot}

\section{Interpretation}\label{sec:interpretation}

\section{Conclusions}

%
%
%
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{paper}

\end{document}

### 2019-04-23 mpc19/paper.tex
types and programs of Pi0 in paper
more delay and elgot

@@
 \usepackage{mathtools}
+\usepackage{fancyvrb}
 
 \newcommand{\N}{\mathbb{N}}
-\newcommand{\Pio}{$\Pi^{\mathsf{o}}$} 
+\newcommand{\Pio}{$\mathsf{\Pi}^{\mathsf{o}}$} 
 \newcommand{\Set}{\mathsf{Set}}
 \newcommand{\comp}{\circ}
+\newcommand{\U}{\mathcal{U}}
+\newcommand{\copair}[2]{[#1,#2]}
+\newcommand{\pair}[2]{\left< #1,#2 \right>}
+\newcommand{\inl}{\mathsf{inl}}
+\newcommand{\inr}{\mathsf{inr}}
+\newcommand{\fst}{\mathsf{fst}}
+\newcommand{\snd}{\mathsf{snd}}
+\newcommand{\T}{\mathsf{T}}
+\newcommand{\Inv}{\mathsf{Inv}}
+\newcommand{\iter}{\mathsf{iter}}
+
+% Syntax Pi0
+\newcommand{\Ty}{\mathsf{Ty}}
+\newcommand{\Var}{\mathsf{Var}}
+\newcommand{\id}{\mathsf{id}}
+\newcommand{\lr}{\longleftrightarrow}
+\newcommand{\fold}{\mathsf{fold}}
+\newcommand{\unfold}{\mathsf{unfold}}
+\newcommand{\sub}{\mathsf{sub}}
+\newcommand{\trace}{\mathsf{trace}}
+\newcommand{\Z}{\mathsf{Z}}
+\newcommand{\I}{\mathsf{I}}
 
@@
 \newcommand{\str}{\mathsf{str}}
+\newcommand{\costr}{\mathsf{costr}}
 \newcommand{\D}{\mathbb{D}}
 \newcommand{\Dapprox}{\mathbb{D}_{\approx}}
+\newcommand{\copairD}[2]{[#1,#2]_{\D}}
+\newcommand{\pairD}[2]{\left< #1,#2 \right>_{\D}}
+\newcommand{\inlD}{\mathsf{inl}_{\D}}
+\newcommand{\inrD}{\mathsf{inr}_{\D}}
+\newcommand{\fstD}{\mathsf{fst}_{\D}}
+\newcommand{\sndD}{\mathsf{snd}_{\D}}
+\newcommand{\piso}{\mathsf{isPartialIso}}
+\newcommand{\pisoalt}{\mathsf{isPartialIso}_2}
+\newcommand{\iterD}{\mathsf{iter}_\D}
 
@@
 %
-\title{Type-theoretic Semantics of \Pio\ via the Delay Monad \thanks{Supported by organization x.}}
+\title{Type-theoretical Semantics of \Pio\ via the Delay Monad \thanks{Supported by organization x.}}
 %
@@
 
+\paragraph{The type-theoretical framework}
+
+Our work is settled in Martin-L\"of type theory with inductive and
+coinductive types. We write $(a : A) \to B \,a$ for dependent function
+spaces and $(a : A) \times B \,a$ for dependent products. We use the
+symbol $=$ for definitional equality of terms and $\equiv$ for
+propositional equality. Given $f : A \to C$ and
+$g : B \to C$, we write $\copair f g : A + B \to C$ for their
+copairing. The coproduct injections are denoted $\inl$ and $\inr$.
+Given $h : C \to A$ and $k : C \to B$, we write
+$\pair h k : C \to A \times B$ for their pairing. The product
+projections are denoted $\fst$ and $\snd$. The empty type is 0 and the
+empty type is 1. We write $\Set$ for the category of
+types and functions between them.
+We define $A \leftrightarrow B = (A \to B) \times (B \to A)$.
+%We write $\U$ for the universe of (small) types.
+
+We do not assume uniqueness of identity proofs
+(UIP), i.e. we do not consider two proofs of $x \equiv y$ necessarily
+equal. Agda natively supports UIP, so we have to manually switch it
+using the \verb|without-K| option.
+
+In Section \ref{sec:delay}, we will need to quotient a certain type by
+an equivalence relation. Martin-L\"of type theory does not support
+quotient types, but quotients can be simulated using setoids
+\cite{BartheCP03}. Alternatively, we can consider extensions of type theory with
+quotient types \`a la Hofmann \cite{Hofmann}, such as homotopy type
+theory \cite{Hott}. Setoids and quotient types \`a la
+Hofmann are not generally equivalent approaches, but they are indeed
+equivalent for the constructions we develop in this work. Therefore,
+in the rest of the paper we assume the existence of quotient types and
+we refrain from technical discussions on their implementation.
+
 \section{Syntax of \Pio}\label{sec:syntax}
+
+\subsection{Types}
+
+\begin{figure}
+\[
+\arraycolsep=8pt\def\arraystretch{2.8}
+\begin{array}{|ccc|}
+\hline 
+\infer{\Z : \Ty_n}{} 
+& \infer{\I : \Ty_n}{}
+& \infer{\Var\,i : \Ty_n}{i \in [0..n)}  \\
+\infer{A \oplus B : \Ty_n}{A : \Ty_n & B : \Ty_n}
+& \infer{A \otimes B : \Ty_n}{A : \Ty_n & B : \Ty_n}
+& \infer{\mu\,A : \Ty_n}{A : \Ty_{n+1}} \\
+\hline
+\end{array}
+\]
+\caption{Types of \Pio.}
+\label{fig:types}
+\end{figure}
+
+\subsection{Programs}
+
+\begin{figure}
+\[
+\arraycolsep=8pt\def\arraystretch{2.8}
+\begin{array}{|cc|}
+\hline
+\infer{\id : A \lr A}{}
+& \infer{g \bullet f : A \lr C}{g : B \lr C & f : A \lr B} \\
+\infer{f \oplus g : A \oplus B \lr C \oplus D}{f : A \lr C & g : B \lr D} 
+& \infer{f \otimes g : A \otimes B \lr C \otimes D}{f : A \lr C & g : B \lr D} \\
+\infer{\lambda_{\oplus} : \Z \oplus A \lr A}{}
+& \infer{\lambda_{\oplus}^{-1} : A \lr \Z \oplus A}{} \\
+\infer{\lambda_{\otimes} : \I \otimes A \lr A}{}
+& \infer{\lambda_{\otimes}^{-1} : A \lr \I \otimes A}{} \\
+\infer{\alpha_{\oplus} : (A \oplus B) \oplus C \lr A \oplus (B \oplus C)}{}
+& \infer{\alpha_{\oplus}^{-1} : A \oplus (B \oplus C) \lr (A \oplus B) \oplus C}{} \\
+\infer{\alpha_{\otimes} : (A \otimes B) \otimes C \lr A \otimes (B \otimes C)}{}
+& \infer{\alpha_{\otimes}^{-1} : A \otimes (B \otimes C) \lr (A \otimes B) \otimes C}{} \\
+\infer{\sigma_{\oplus} : A \oplus B \lr B \oplus A}{}
+& \infer{\sigma_{\otimes} : A \otimes B \lr B \otimes A}{} \\
+\infer{\kappa : \Z \otimes A \lr \Z}{} 
+& \infer{\delta : (A \oplus B) \otimes C \lr (A \otimes C) \oplus (B \otimes C)}{} \\
+\infer{\kappa^{-1} : \Z \lr \Z \otimes A}{}
+& \infer{\delta^{-1} : (A \otimes C) \oplus (B \otimes C)  \lr (A \oplus B) \otimes C}{} \\
+\infer{\fold : \sub \, A \, (\mu \,A) \lr \mu \,A}{} 
+& \infer{\unfold : \mu\,A \lr \sub \, A \, (\mu \,A)}{} \\
+\multicolumn{2}{|c|}{\infer{\trace \,f: B \lr C}{f : A \oplus B \lr A \oplus C}} \\
+\hline
+\end{array}
+\]
+\caption{Programs of \Pio}
+\label{fig:programs}
+\end{figure}
 
@@
 \[
-\infer={\now\,a : \Delay A}{x : A}
+\infer={\now\,a : \Delay A}{a : A}
 \quad
@@
 \]
-(Here and in the following, single rule lines refer to an inductive
-definition, double rule lines to a coinductive definition.) The
+(Double rule lines refer to a coinductive definition.) The
 constructor $\now$ embeds $A$ into $\Delay A$, so $\now\,a$ represents
@@
 operation which we denote by $\str : A \times \Delay B \to \Delay (A
-\times B)$. Moreover, the delay datatype is a commutative monad.
+\times B)$. Similarly, it has a unique costrength operation $\costr :
+\Delay A \times B \to \Delay (A \times B)$ definable using $\str$. Moreover, the delay datatype is a commutative monad.
 
@@
 partiality, which in particular allows the specification of a
-well-behaved trace operator, we need to quotient the delay datatype by
+well-behaved trace operator, we introduce the notion of termination-sensitive
 weak bisimilarity.
@@
 informal statement can be formalized in several different but
-logically equivalent ways \cite{ChapmanUV19}. For this work we chose the
+logically equivalent ways \cite{ChapmanUV19}. Here we consider the
 following definition:
@@
 The definition says that $x$ and $y$ are weakly bisimilar if, whenever
-$x$ terminates, then $y$ also terminates with the same value, and vice
-versa. Weak bisimilarity is an equivalence relation and it is a
-congrence with respect to the $\later$ operation.  We call $\Dapprox$
-the Kleisli category of $\Delay$ with homsets quotiented by pointwise
-weak bisimilarity. This means that in $\Dapprox$ two morphisms $f$ and
-$g$ are considered equal whenever $f \, a \approx g \, a$, for all
-inputs $a$. When this is the case, we also write $f \approx g$.  In
-this work we do not formally construct the quotients
-$(A \to \Delay B)/{\approx}$, since we work in a type theory without
-quotient types. Instead we require that every construction we perform
-on morphisms in $\Dapprox$ respects weak bisimilarity. Technically, we
-construct $\Dapprox$ as a category enriched in setoids CITE, where in
-particular all the reasoning on morphisms is performed up-to weak
-bisimilarity.
+$x$ terminates returning a value $a$, then $y$ also terminates
+returning $a$, and vice versa. Weak bisimilarity is an equivalence
+relation and it is a congruence w.r.t. the $\later$ operation.
+We call $\Dapprox$ the category $\D$ with homsets
+quotiented by pointwise weak bisimilarity. This means that in
+$\Dapprox$ two morphisms $f$ and $g$ are considered equal whenever
+$f \, a \approx g \, a$, for all inputs $a$. When this is the case, we
+also write $f \approx g$.
 
-In previous work \cite{ChapmanUV19} we showed that, when working in a
-type theory with quotient types a la Hofmann CITE, quotienting the
-delay monad by weak bisimilarity has problematic consequences. In
-fact, it does not seem possible to recover the monad structure on the
-quotiented datatype $\mathsf{DelayQuot}\,A = \Delay A/{\approx}$
-without assuming additional controversial principles
-such as the axiom of countable choice. In particolar, countable choice
-would be necessary for modelling the trace operator of \Pio. In
-loc.cit. we showed that a possible way to avoid these issues is
-forming the quotients of homsets, as in $(A \to \Delay B)/{\approx}$, instead of
-quotienting the delay datatype directly. We would still not be able to form
+As an alternative to quotienting the homsets of $\D$, we could have
+quotiented the delay datatype by weak bisimilarity:
+$\mathsf{DelayQuot}\,A = \Delay A/{\approx}$. In previous work
+\cite{ChapmanUV19}, we showed that this construction has problematic
+consequences if we employ Hofmann's approach to quotient types
+\cite{Hofmann}. For example, it does not seem possible to lift the
+monad structure of $\Delay$ to $\mathsf{DelayQuot}$ without
+postulating additional principles such as the axiom of countable
+choice. More fundamentally for this work, countable choice would be
+needed for modelling the trace operator of \Pio\ in the Kleisli
+category of $\mathsf{DelayQuot}$. Notice that, if the setoid approach
+to quotienting is employed, the latter constructions go through
+without the need for additional assumptions. In order to keep an
+agnostic perspective on quotient types and avoid the need for
+disputable semi-classical choice principles, we decided to quotient
+the homsets of $\D$ by (poitwise) weak bisimilarity instead of the
+delay datatype.
+
+\subsection{Finite Products and Coproducts}
+\label{sec:prod}
+
+Colimits in $\Dapprox$ are inherited from $\Set$. This means that 0 is
+also the initial object of $\Dapprox$, similarly $A + B$ is the
+binary coproduct of $A$ and $B$ in $\Dapprox$. Given $f : A \to
+\Delay C$ and $g : B \to \Delay C$, we define their copairing as
+$\copairD f g = \copair f g : A + B \to \Delay C$. The operation
+$\copairD - -$ is compatible with weak bisimilarity, in the sense that
+$\copairD{f_1}{g_1} \approx \copairD{f_2}{g_2}$ whenever $f_1 \approx f_2$
+and $g_1 \approx g_2$.
+The coproduct
+injections are given by $\inlD = \now \comp \inl : A \to \Delay (A +
+B)$ and $\inrD = \now \comp \inr : B \to \Delay (A + B)$.
+
+Limits in $\Set$ do not lift to $\Dapprox$. In particular, 1 is not
+terminal object and $A \times B$ is not the binary product of $A$ and
+$B$ in $\Dapprox$. In fact, 0 is the terminal object, with
+$\lambda \_.\, \never : A \to \Delay 0$ as the terminal morphism.  The
+binary product of $A$ and $B$ in $\Dapprox$ is given by
+$A + B + A \times B$. Nevertheless, it is possible to prove that 1 and
+$- \times -$ are partial terminal object and partial binary products
+respectively, in the sense of Cockett and Lack's restriction
+categories \cite{CockettL07}. Here we refrain from making the latter
+statement formal. We only show the construction of the partial pairing
+operation, which we employ in the interpretation of \Pio.  Given
+$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
+\begin{align*}
+& \pairD f g : C \to \Delay (A \times B) \\
+& \pairD f g = \costr \diamond (\str \comp \pair f g)
+\end{align*}
+Since the delay monad is commutative, the function $\pairD f g$ is
+equal to $\str \diamond (\costr \comp \pair f g)$. The operation
+$\pairD - -$ is compatible with weak bisimilarity.
+
+\subsection{Partial Isomorphisms}
+\label{sec:isos}
+
+In order to model the reversible programs of \Pio, we need to consider
+reversible computations in $\Dapprox$. Given a morphism $f : A \to
+\Delay B$, we say that it is a partial isomorphism if the following
+type is inhabited:
+\begin{align*}
+\piso \,f = (g : B \to \Delay A) \times \left( (a : A) (b : B) \to  f
+  \,a \dn b \leftrightarrow g \,b \dn a  \right)  
+\end{align*}
+In other words, $f$ is a partial isomorphism if there exists a
+morphism $g : B \to \Delay A$ such that, if $f\,a$ terminates
+returning a value $b$, then $g\,b$ terminates returning $a$, and vice
+versa. 
+Our definition of partial isomorhisms is equivalent to the standard
+categorical one CITATION, which translated in our type-theoretical
+setting becomes
+\begin{align*}
+\pisoalt \,f = (g : B \to \Delay A) \times f \diamond g \diamond
+  f \approx f \times g \diamond f \diamond g \approx g 
+\end{align*}
+
+We denote $A \simeq B$ the type of partial isomorphisms between
+$A$ and $B$:
+\[
+A \simeq B = (f : A \to \Delay B) \times \piso \,f
+\]
+We call $\Inv \Dapprox$ the subcategory of $\Dapprox$ consisting
+of (equivalence classes of) partial isomorphisms. The category
+$\Inv\Dapprox$ inherits partial finite products from $\Dapprox$, but
+not the coproducts. It is possible to prove though that, in the category
+$\Inv\Dapprox$, 0 is a partial zero and $A + B$ is the disjointness
+tensor of $A$ and $B$ \cite{Giles}.
+
+%%  In this work we do not formally construct
+%% the quotients $(A \to \Delay B)/{\approx}$, since we work in a type
+%% theory without quotient types. 
+%% % Instead we require that every construction we perform on morphisms in
+%% % $\Dapprox$ respects weak bisimilarity.
+%% Instead, we represent the homsets of $\Dapprox$ as setoids.a category
+%% enriched in setoids \cite{BartheCP03}: $\Dapprox$ has the same objects
+%% of $\D$, but morphisms between $A$ and $B$ form a setoid $(A \to
+%% \Delay B, {\approx})$. This implies that every function we
+%% define out of the type $A \to \Delay B$ must respects weak
+%% bisimilarity.
+
+%% In previous work \cite{ChapmanUV19} we showed that, when working in a
+%% type theory with quotient types a la Hofmann \cite{Hofmann}, quotienting the
+%% delay monad by weak bisimilarity has problematic consequences. For example,
+%% it does not seem possible to recover the monad structure on the
+%% quotiented datatype $\mathsf{DelayQuot}\,A = \Delay A/{\approx}$
+%% without assuming non-constructive principles
+%% such as the axiom of countable choice. More fundamentally for this
+%% work, countable choice would be needed for modelling the trace operator of \Pio. In
+%% loc.cit. we showed that a possible way to avoid these issues is
+%% forming the quotients of homsets, as in $(A \to \Delay B)/{\approx}$, instead of
+%% quotienting the delay datatype directly.
 
 \section{Elgot Iteration and Trace}\label{sec:elgot}
+
+An Elgot monad is a monad $T$ whose Kleilsi category supports
+unguarded Elgot iteration. More precisely\footnote{Here we give the
+  definition of Elgot monad on $\Set$, but the definition of Elgot
+  monad makes sense in any category with binary coproducts.}, a monad
+$\T$ is Elgot if there exists an operation
+\[
+\iter_\T : (A \to \T \,(A + B)) \to A \to \T\,B
+\]
+satisfying a number of axioms, most notably the unfolding axiom:
+\[
+\iter_\T\,f \equiv \copair{\iter_\T \,f}{\eta_\T} \diamond_\T f
+\]
+where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
+composition in the Kleisli category of $\T$.
+
+The delay monad is an Elgot monad up to weak bisimilarity. Its
+iteration operator relies on the definition of an auxiliary function
+$\iterD'$ corecursively defined as follows:
+\begin{align*}
+& \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
+& \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f a)) \\
+& \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
+& \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
+& \\
+& \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
+& \iterD \,f \,a = \iterD' \,f \,(f \,a)
+\end{align*}
+The definition above can be given the following intuitive
+explaination. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
+not terminate as well. If $f\,a$ terminates, there are two
+possibilities: either $f\,a$ converges to $\inr\,b$, in which case
+$\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
+to $\inl\,a'$, in which case we repeat the precedure by replacing $a$
+with $a'$. Notice that in the latter case we also add one occurrence
+of $\later$ to the total computation time. This addition is necessary
+achieving the productivity of the corecursively defined function
+$\iterD'$. In fact, by changing the first line of its specification to
+$\iterD' \,f \,(\now \,(\inl\, a)) = \iterD' \,f \,(f a)$
+and taking $f = \inlD$, we would have that $\iterD' \,f \,(\now
+\,(\inl\, a))$ unfolds indefinitely without producing any output. In
+Agda, such a definition would be rightfully rejected by the termination
+checker.
 

### 2019-04-24 mpc19/paper.tex
alternative weak bisim

@@
 \[
-\infer={\now\,a : \Delay A}{a : A}
+\infer{\now\,a : \Delay A}{a : A}
 \quad
@@
 \]
-(Double rule lines refer to a coinductive definition.) The
+(Double rule lines refer to a coinductive constructor.) The
 constructor $\now$ embeds $A$ into $\Delay A$, so $\now\,a$ represents
@@
 informal statement can be formalized in several different but
-logically equivalent ways \cite{ChapmanUV19}. Here we consider the
-following definition:
-\[
-x \approx y = (a : A) \to (x \dn a \to y \dn a) \times (y \dn a \to x \dn a) 
-\]
+logically equivalent ways \cite{ChapmanUV19}. 
+A possible defintion is as follows:
+\begin{equation}\label{eq:weak1}
+x \approx y = (a : A) \to x \dn a \leftrightarrow y \dn a
+\end{equation}
 The definition says that $x$ and $y$ are weakly bisimilar if, whenever
 $x$ terminates returning a value $a$, then $y$ also terminates
-returning $a$, and vice versa. Weak bisimilarity is an equivalence
+returning $a$, and vice versa.
+Equivalently, weak bisimilarity can be coinductively formulated as follows:
+\begin{equation}
+\label{eq:weak2}
+\infer{\now\,a \approx \now\,a}{}
+\quad
+\infer{\now\,a \approx \later\,x}{\now\,a \approx x}
+\quad
+\infer{\later\,x \approx \now\,a}{x \approx \now \,a}
+\quad
+\infer={\later\,x_1 \approx \later\,x_2}{x_1 \approx x_2}
+\end{equation}
+In the rest of the paper we will employ
+both encodings of weak bisimilarity, in each case explicitly stating
+if we are working with definition (\ref{eq:weak1}) or definition (\ref{eq:weak2}).
+Weak bisimilarity is an equivalence
 relation and it is a congruence w.r.t. the $\later$ operation.
@@
 quotiented the delay datatype by weak bisimilarity:
-$\mathsf{DelayQuot}\,A = \Delay A/{\approx}$. In previous work
+$\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$. In previous work
 \cite{ChapmanUV19}, we showed that this construction has problematic
@@
 \cite{Hofmann}. For example, it does not seem possible to lift the
-monad structure of $\Delay$ to $\mathsf{DelayQuot}$ without
+monad structure of $\Delay$ to $\mathsf{Delay}_{\approx}$ without
 postulating additional principles such as the axiom of countable
@@
 needed for modelling the trace operator of \Pio\ in the Kleisli
-category of $\mathsf{DelayQuot}$. Notice that, if the setoid approach
+category of $\mathsf{Delay}_{\approx}$. Notice that, if the setoid approach
 to quotienting is employed, the latter constructions go through
@@
 
-The delay monad is an Elgot monad up to weak bisimilarity. Its
-iteration operator relies on the definition of an auxiliary function
-$\iterD'$ corecursively defined as follows:
+The delay monad is an Elgot monad for which the axioms holds up to
+weak bisimilarity, not propositional equality. Its iteration operator
+relies on the definition of an auxiliary function $\iterD'$
+corecursively defined as follows:
 \begin{align*}
 & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
-& \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f a)) \\
+& \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
 & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\

### 2019-04-24 mpc19/paper.tex
Added some TODOs!

@@
 \usepackage{fancyvrb}
+
+\usepackage{soul}
 
@@
 
+\newcommand{\todo}[1]{\hl{TODO: #1}}
+\newcommand{\todocite}[1]{\hl{[TODO: #1]}}
+
 \begin{document}
@@
 
+\section{TODO} % (fold)
+\label{sec:todo}
+Robin:
+\begin{itemize}
+  \item Abstract.
+  \item Introduction.
+  \item Check and correct 4.1, 4.2, add citations, etc.
+  \item Make pictures! 
+  \begin{itemize}
+    \item Graphical language of monoidal categories, and for \Pio.
+    \item Trace.
+    \item Iterator, iterator unfolding axiom.
+    \item Trace from iterator.
+  \end{itemize}
+  \item Conclusion
+  \begin{itemize}
+    \item Related work: Kleisli category of the Delay monad is a join
+    restriction category, inverse category of a join restriction category is a
+    join inverse category; already known to be dagger-traced under certain
+    conditions.
+    \item Future work: Semantics of time invertible programming. Extension of
+    the work on the Kleisli category of the Delay monad beyond $\Set$ to other
+    (more general) categories, e.g. topoi.
+  \end{itemize}
+\end{itemize}
+Niccolò: 
+\begin{itemize}
+  \item Fix the syntax to be closer to ordinary presentations (but do describe 
+  how it is realized in Agda).
+  \item Finish the section about trace and iteration.
+  \item Remind the reader about double lines of coinductive definitions. 
+\end{itemize}
+Both:
+\begin{itemize}
+  \item \textbf{Important:} Come up with a punnier title! (Something about 
+  delays and daggers).
+\end{itemize}
+% section todo (end)
+
 \section{Introduction}\label{sec:intro}
+\todo{RC Motivation.}
+
+\todo{Formalization motivation? Maybe?}
+
+\todo{Explain Carette et al. Say that we are extending it to be actually useful (but don't say it like that).}
+
+\todo{Delay vs. Maybe monad.}
 

### 2019-04-24 mpc19/paper.tex
Fixed 4.1, 4.2 and added citations.

@@
 \newcommand{\Set}{\mathsf{Set}}
+\newcommand{\Par}{\mathsf{Par}}
 \newcommand{\comp}{\circ}
@@
   \item Introduction.
-  \item Check and correct 4.1, 4.2, add citations, etc.
   \item Make pictures! 
@@
 
-Limits in $\Set$ do not lift to $\Dapprox$. In particular, 1 is not
-terminal object and $A \times B$ is not the binary product of $A$ and
-$B$ in $\Dapprox$. In fact, 0 is the terminal object, with
-$\lambda \_.\, \never : A \to \Delay 0$ as the terminal morphism.  The
+Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
+and partial functions, they do not lift to $\Dapprox$ either. This is not an
+issue with these concrete formulations of partiality, but rather with the
+interaction of partiality (in the sense of restriction categories) and limits
+in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is not
+terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
+$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
+\never : A \to \Delay 0$ as the terminal morphism. The
 binary product of $A$ and $B$ in $\Dapprox$ is given by
@@
 respectively, in the sense of Cockett and Lack's restriction
-categories \cite{CockettL07}. Here we refrain from making the latter
+categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
 statement formal. We only show the construction of the partial pairing
@@
 returning a value $b$, then $g\,b$ terminates returning $a$, and vice
-versa. 
-Our definition of partial isomorhisms is equivalent to the standard
-categorical one CITATION, which translated in our type-theoretical
-setting becomes
+versa. In $\Dapprox$, our definition of partial isomorphisms is equivalent to
+the standard categorical one~\cite{Kastl79} (see also \cite{CockettL02}),
+which, translated in our type-theoretical setting, is
 \begin{align*}
@@
 \end{align*}
-
 We denote $A \simeq B$ the type of partial isomorphisms between
@@
 \]
-We call $\Inv \Dapprox$ the subcategory of $\Dapprox$ consisting
-of (equivalence classes of) partial isomorphisms. The category
-$\Inv\Dapprox$ inherits partial finite products from $\Dapprox$, but
-not the coproducts. It is possible to prove though that, in the category
-$\Inv\Dapprox$, 0 is a partial zero and $A + B$ is the disjointness
-tensor of $A$ and $B$ \cite{Giles}.
+We call $\Inv \Dapprox$ the subcategory of $\Dapprox$ consisting of
+(equivalence classes of) partial isomorphisms. Note that $\Inv\Dapprox$
+inherits neither partial products nor coproducts of $\Dapprox$, as the
+universal mapping property fails in both cases. However, it can be shown that
+in the category $\Inv\Dapprox$, $0$ is a zero object, $A + B$ is the
+disjointness tensor product of $A$ and $B$ (in the sense of Giles~\cite{Giles})
+with unit $0$, and $A \times B$ a monoidal product of $A$ and $B$ with unit $1$
+(though it is \emph{not} an inverse product in the sense of
+Giles~\cite{Giles}, as that would imply decidable equality on all objects).
 

### 2019-04-24 mpc19/paper.tex
proof of unfolding axiom

--- 2019-04-24 mpc19/paper.tex
@@
 \newcommand{\later}{\mathsf{later}}
+\newcommand{\laterR}{\mathsf{laterR}}
+\newcommand{\laterL}{\mathsf{laterL}}
 \newcommand{\never}{\mathsf{never}}
@@
 \newcommand{\iterD}{\mathsf{iter}_\D}
+\newcommand{\unfolding}{\mathsf{unfolding}}
+\newcommand{\refl}{\mathsf{refl}}
 
@@
 coinductive types. We write $(a : A) \to B \,a$ for dependent function
-spaces and $(a : A) \times B \,a$ for dependent products. We use the
+spaces and $(a : A) \times B \,a$ for dependent products. We allow
+dependent functions to have implicit arguments and indicate implicit
+argument positions with curly brackets (as in Agda). We use the
 symbol $=$ for definitional equality of terms and $\equiv$ for
@@
 
-\begin{figure}
 \[
-\arraycolsep=8pt\def\arraystretch{2.8}
-\begin{array}{|ccc|}
-\hline 
-\infer{\Z : \Ty_n}{} 
-& \infer{\I : \Ty_n}{}
-& \infer{\Var\,i : \Ty_n}{i \in [0..n)}  \\
-\infer{A \oplus B : \Ty_n}{A : \Ty_n & B : \Ty_n}
-& \infer{A \otimes B : \Ty_n}{A : \Ty_n & B : \Ty_n}
-& \infer{\mu\,A : \Ty_n}{A : \Ty_{n+1}} \\
-\hline
-\end{array}
+A ::= \Z \, | \,A \oplus A \, | \,\I \,| \,A \otimes A \,| \,X \,|
+\,\mu X.\,A 
 \]
-\caption{Types of \Pio.}
-\label{fig:types}
-\end{figure}
+
+%% \begin{figure}
+%% \[
+%% \arraycolsep=8pt\def\arraystretch{2.8}
+%% \begin{array}{|ccc|}
+%% \hline 
+%% \infer{\Z : \Ty_n}{} 
+%% & \infer{\I : \Ty_n}{}
+%% & \infer{\Var\,i : \Ty_n}{i \in [0..n)}  \\
+%% \infer{A \oplus B : \Ty_n}{A : \Ty_n & B : \Ty_n}
+%% & \infer{A \otimes B : \Ty_n}{A : \Ty_n & B : \Ty_n}
+%% & \infer{\mu\,A : \Ty_n}{A : \Ty_{n+1}} \\
+%% \hline
+%% \end{array}
+%% \]
+%% \caption{Types of \Pio.}
+%% \label{fig:types}
+%% \end{figure}
 
@@
 \label{eq:weak2}
-\infer{\now\,a \approx \now\,a}{}
-\quad
-\infer{\now\,a \approx \later\,x}{\now\,a \approx x}
-\quad
-\infer{\later\,x \approx \now\,a}{x \approx \now \,a}
-\quad
-\infer={\later\,x_1 \approx \later\,x_2}{x_1 \approx x_2}
+\arraycolsep=8pt\def\arraystretch{2.8}
+\begin{array}{cc}
+\infer{\now_{\approx}  : \now\,a \approx \now\,a}{}
+&
+\infer={\later_{\approx} \,p : \later\,x_1 \approx \later\,x_2}{p : x_1 \approx x_2}
+\\
+\infer{\laterL_{\approx} \, p: \later\,x \approx \now\,a}{p : x \approx \now \,a}
+&
+\infer{\laterR_{\approx} \, p: \now\,a \approx \later\,x}{p : \now\,a \approx x}
+\end{array}
 \end{equation}
@@
 relation and it is a congruence w.r.t. the $\later$ operation.
+For example, here is a proof that weak bisimilarity is reflexive,
+employing definition (\ref{eq:weak2}).
+\begin{align*}
+& \refl_{\approx} : \{x : \Delay A\} \to x \approx x \\
+& \refl_{\approx} \,\{\now\,a\} = \now_\approx \\
+& \refl_{\approx} \,\{\later\,x\} = \later_\approx\,(\refl_\approx \,\{x\})
+\end{align*}
+
 We call $\Dapprox$ the category $\D$ with homsets
@@
 corecursively defined as follows:
-\begin{align*}
-& \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
-& \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
-& \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
-& \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
-& \\
-& \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
-& \iterD \,f \,a = \iterD' \,f \,(f \,a)
-\end{align*}
+\[ %\arraycolsep=8pt
+\def\arraystretch{1.2}
+\begin{array}{ll}
+\multicolumn{2}{l}{\iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B} \\
+\iterD' \,f \,(\now \,(\inl\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
+\iterD' \,f \,(\now \,(\inr\, b)) &= \now \,b \\
+\iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
+ \\
+\multicolumn{2}{l}{\iterD : (A \to \Delay (A + B)) \to A \to \Delay B} \\
+\multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
+\end{array}
+\]
+
+%% \begin{align*}
+%% & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
+%% & \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
+%% & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
+%% & \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
+%% & \\
+%% & \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
+%% & \iterD \,f \,a = \iterD' \,f \,(f \,a)
+%% \end{align*}
 The definition above can be given the following intuitive
@@
 checker.
+
+As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
+up to weak bisimilarity. Here we show the proof of the unfolding
+axiom. We employ the definition (\ref{eq:weak2}) of weak
+bisimilarity. The proof relies on an auxiliary proof $\unfolding'$. 
+\[
+\def\arraystretch{1.2}
+\begin{array}{ll}
+\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (A + B)) \to
+  \bind\, \copairD{\iterD \,f}{\now} \approx \iterD' \,f} \\
+\unfolding' \,f \,(\now \,(\inl \,a)) & = \laterR_{\approx} \,
+  \refl_\approx \\
+\unfolding' \,f \,(\now \,(\inr \,b)) & = \now_\approx \\
+\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
+  x) \\
+ \\
+\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (A + B)) \to \copairD{\iterD
+  \,f}{\now} \diamond f \approx \iterD \,f}\\
+\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
+\end{array}
+\]
 

### 2019-04-24 mpc19/paper.tex
Added TikZiT support, basic diagrams.
finished unfolding

@@
 \]
-(Double rule lines refer to a coinductive constructor.) The
-constructor $\now$ embeds $A$ into $\Delay A$, so $\now\,a$ represents
-the terminating computation returning the value $a$. The constructor
-$\later$ adds an additional unit of time delay to a computation. The
-non-terminating computation $\never$ is corecursively defined as
-$\never = \later \,\never$.
+The constructor $\now$ embeds $A$ into $\Delay A$, so $\now\,a$
+represents the terminating computation returning the value $a$. The
+constructor $\later$ adds an additional unit of time delay to a
+computation. Double rule lines refer to a coinductive constructor,
+which can be applied an infinite amount of times.
+The non-terminating computation $\never$
+is corecursively defined as $\never = \later \,\never$.
 
@@
 returning $a$, and vice versa.
-Equivalently, weak bisimilarity can be coinductively formulated as follows:
+Alternatively, weak bisimilarity can be coinductively formulated as follows:
 \begin{equation}
@@
 \end{equation}
+Notice that the constructor $\later_{\approx}$ is
+coinductive. Formulation (\ref{eq:weak1}) and (\ref{eq:weak2}) are
+logically equivalent, meaning that two computations are weakly
+bisimilar as in (\ref{eq:weak1}) if and only if they are weakly
+bisimilar as in (\ref{eq:weak2}).
 In the rest of the paper we will employ
@@
 
-\section{Elgot Iteration and Trace}\label{sec:elgot}
+\section{Elgot Iteration}\label{sec:elgot}
 
@@
 
+\subsection{Trace}\label{sec:trace}
+
+\subsection{Dagger Trace}\label{sec:daggertrace}
+
 \section{Interpretation}\label{sec:interpretation}

### 2019-04-24 mpc19/paper.tex
Added diagrams for the construction of the trace from the Elgot iterator.

--- 2019-04-24 mpc19/paper.tex
@@
 \usepackage{fancyvrb}
+
+\usepackage{tikzit}
+\input{figures/stringdiagrams.tikzstyles}
 
@@
 \newcommand{\later}{\mathsf{later}}
-\newcommand{\laterR}{\mathsf{laterR}}
-\newcommand{\laterL}{\mathsf{laterL}}
 \newcommand{\never}{\mathsf{never}}
@@
 \newcommand{\iterD}{\mathsf{iter}_\D}
-\newcommand{\unfolding}{\mathsf{unfolding}}
-\newcommand{\refl}{\mathsf{refl}}
 
@@
 coinductive types. We write $(a : A) \to B \,a$ for dependent function
-spaces and $(a : A) \times B \,a$ for dependent products. We allow
-dependent functions to have implicit arguments and indicate implicit
-argument positions with curly brackets (as in Agda). We use the
+spaces and $(a : A) \times B \,a$ for dependent products. We use the
 symbol $=$ for definitional equality of terms and $\equiv$ for
@@
 
+\begin{figure}
 \[
-A ::= \Z \, | \,A \oplus A \, | \,\I \,| \,A \otimes A \,| \,X \,|
-\,\mu X.\,A 
+\arraycolsep=8pt\def\arraystretch{2.8}
+\begin{array}{|ccc|}
+\hline 
+\infer{\Z : \Ty_n}{} 
+& \infer{\I : \Ty_n}{}
+& \infer{\Var\,i : \Ty_n}{i \in [0..n)}  \\
+\infer{A \oplus B : \Ty_n}{A : \Ty_n & B : \Ty_n}
+& \infer{A \otimes B : \Ty_n}{A : \Ty_n & B : \Ty_n}
+& \infer{\mu\,A : \Ty_n}{A : \Ty_{n+1}} \\
+\hline
+\end{array}
 \]
-
-%% \begin{figure}
-%% \[
-%% \arraycolsep=8pt\def\arraystretch{2.8}
-%% \begin{array}{|ccc|}
-%% \hline 
-%% \infer{\Z : \Ty_n}{} 
-%% & \infer{\I : \Ty_n}{}
-%% & \infer{\Var\,i : \Ty_n}{i \in [0..n)}  \\
-%% \infer{A \oplus B : \Ty_n}{A : \Ty_n & B : \Ty_n}
-%% & \infer{A \otimes B : \Ty_n}{A : \Ty_n & B : \Ty_n}
-%% & \infer{\mu\,A : \Ty_n}{A : \Ty_{n+1}} \\
-%% \hline
-%% \end{array}
-%% \]
-%% \caption{Types of \Pio.}
-%% \label{fig:types}
-%% \end{figure}
+\caption{Types of \Pio.}
+\label{fig:types}
+\end{figure}
 
@@
 \label{eq:weak2}
-\arraycolsep=8pt\def\arraystretch{2.8}
-\begin{array}{cc}
-\infer{\now_{\approx}  : \now\,a \approx \now\,a}{}
-&
-\infer={\later_{\approx} \,p : \later\,x_1 \approx \later\,x_2}{p : x_1 \approx x_2}
-\\
-\infer{\laterL_{\approx} \, p: \later\,x \approx \now\,a}{p : x \approx \now \,a}
-&
-\infer{\laterR_{\approx} \, p: \now\,a \approx \later\,x}{p : \now\,a \approx x}
-\end{array}
+\infer{\now\,a \approx \now\,a}{}
+\quad
+\infer{\now\,a \approx \later\,x}{\now\,a \approx x}
+\quad
+\infer{\later\,x \approx \now\,a}{x \approx \now \,a}
+\quad
+\infer={\later\,x_1 \approx \later\,x_2}{x_1 \approx x_2}
 \end{equation}
@@
 relation and it is a congruence w.r.t. the $\later$ operation.
-For example, here is a proof that weak bisimilarity is reflexive,
-employing definition (\ref{eq:weak2}).
-\begin{align*}
-& \refl_{\approx} : \{x : \Delay A\} \to x \approx x \\
-& \refl_{\approx} \,\{\now\,a\} = \now_\approx \\
-& \refl_{\approx} \,\{\later\,x\} = \later_\approx\,(\refl_\approx \,\{x\})
-\end{align*}
-
 We call $\Dapprox$ the category $\D$ with homsets
@@
 corecursively defined as follows:
-\[ %\arraycolsep=8pt
-\def\arraystretch{1.2}
-\begin{array}{ll}
-\multicolumn{2}{l}{\iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B} \\
-\iterD' \,f \,(\now \,(\inl\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
-\iterD' \,f \,(\now \,(\inr\, b)) &= \now \,b \\
-\iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
- \\
-\multicolumn{2}{l}{\iterD : (A \to \Delay (A + B)) \to A \to \Delay B} \\
-\multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
-\end{array}
-\]
-
-%% \begin{align*}
-%% & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
-%% & \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
-%% & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
-%% & \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
-%% & \\
-%% & \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
-%% & \iterD \,f \,a = \iterD' \,f \,(f \,a)
-%% \end{align*}
+\begin{align*}
+& \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
+& \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
+& \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
+& \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
+& \\
+& \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
+& \iterD \,f \,a = \iterD' \,f \,(f \,a)
+\end{align*}
 The definition above can be given the following intuitive
@@
 
-As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
-up to weak bisimilarity. Here we show the proof of the unfolding
-axiom. We employ the definition (\ref{eq:weak2}) of weak
-bisimilarity. The proof relies on an auxiliary proof $\unfolding'$. 
-\[
-\def\arraystretch{1.2}
-\begin{array}{ll}
-\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (A + B)) \to
-  \bind\, \copairD{\iterD \,f}{\now} \approx \iterD' \,f} \\
-\unfolding' \,f \,(\now \,(\inl \,a)) & = \laterR_{\approx} \,
-  \refl_\approx \\
-\unfolding' \,f \,(\now \,(\inr \,b)) & = \now_\approx \\
-\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
-  x) \\
- \\
-\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (A + B)) \to \copairD{\iterD
-  \,f}{\now} \diamond f \approx \iterD \,f}\\
-\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
-\end{array}
-\]
+\ctikzfig{fL_fR}
+\ctikzfig{trace}
 

### 2019-04-24 mpc19/paper.tex
Hasegaws

--- 2019-04-24 mpc19/paper.tex
@@
 
+\usepackage{tikzit}
+\input{figures/stringdiagrams.tikzstyles}
+
 \usepackage{soul}
@@
 \newcommand{\iter}{\mathsf{iter}}
+\newcommand{\mapL}[1]{#1_{\mathsf{L}}}
+\newcommand{\mapR}[1]{#1_{\mathsf{R}}}
 
@@
 \newcommand{\refl}{\mathsf{refl}}
+\newcommand{\traceD}{\mathsf{trace}_\D}
+\newcommand{\traceHD}{\mathsf{traceH}_\D}
 
@@
 
+The operation $\iterD$ is compatible with weak bisimilarity, which means
+that $\iterD\,f_1 \approx \iterD\,f_2$ whenever $f_1 \approx f_2$.
+
 As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
@@
 
+From the Elgot iteration operator it is possible to derive a trace
+operator. First, given $f : A + B \to C$, we introduce
+$\mapL f = f \comp \inl : A \to C$ and
+$\mapR f = f \comp \inr : B \to C$, so that $f = [\mapL f , \mapR f]$.
+The trace operator in $\Dapprox$ is defined as follows:
+\begin{align*}
+& \traceD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
+& \traceD \,f = \copairD{\iterD \,\mapL f}{\now} \diamond \mapR f
+\end{align*}
+The operation $\traceD$ is compatible with weak bisimilarity.
+
+It is well-known that a trace operator is obtainable from an iteration
+operator. This is how Hasegawa's construction  \cite{Hasegwa97} derives it in the following way:
+\begin{align*}
+& \traceD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
+& \traceD \,f = \copairD{\iterD \,\mapL f}{\now} \diamond \mapR f
+\end{align*}
+
 \subsection{Dagger Trace}\label{sec:daggertrace}
+
+%\ctikzfig{language}
 

### 2019-04-24 mpc19/paper.tex

@@
 
+\ctikzfig{fL_fR}
+\ctikzfig{trace}
+
 It is well-known that a trace operator is obtainable from an iteration
-operator. This is how Hasegawa's construction  \cite{Hasegwa97} derives it in the following way:
+operator, as shown by Hasegawa  \cite{Hasegawa97}. His construction,
+instantiated to our setting, looks like this:
 \begin{align*}
-& \traceD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
-& \traceD \,f = \copairD{\iterD \,\mapL f}{\now} \diamond \mapR f
+& \traceHD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
+& \traceHD \,f = \iterD (\Delay (\inl + \id) \comp f) \comp \inr
 \end{align*}
@@
 \subsection{Dagger Trace}\label{sec:daggertrace}
-
-%\ctikzfig{language}
 

### 2019-04-25 mpc19/paper.tex
have to change order of types in iter
Updated TODO.
More TODO stuff!
Added a bunch of references, introduction.

@@
 \newcommand{\N}{\mathbb{N}}
-\newcommand{\Pio}{$\mathsf{\Pi}^{\mathsf{o}}$} 
+\newcommand{\Pio}{\ensuremath{\mathsf{\Pi}^{\mathsf{o}}}} 
 \newcommand{\Set}{\mathsf{Set}}
@@
 \newcommand{\sub}{\mathsf{sub}}
-\newcommand{\trace}{\mathsf{trace}}
+\newcommand{\trace}{\ensuremath{\mathsf{trace}}}
 \newcommand{\Z}{\mathsf{Z}}
@@
 % Delay macros
-\newcommand{\Delay}{\mathsf{Delay}\,}
+\newcommand{\Delay}{\ensuremath{\mathsf{Delay}\,}}
 \newcommand{\now}{\mathsf{now}}
@@
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
+\todo{Write it!}
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+\keywords{reversible computation \and iteration \and delay monad}
 \end{abstract}
@@
   \begin{itemize}
-    \item Graphical language of monoidal categories, and for \Pio.
-    \item Trace.  
-    \item Iterator, iterator unfolding axiom.
-    \item Trace from iterator.
+    \item Hasegawa trace from iterator.
   \end{itemize}
@@
   \item Remind the reader about double lines of coinductive definitions.
+  \item Swap types in Elgot iterator.
 \end{itemize}
@@
 \section{Introduction}\label{sec:intro}
-\todo{RC Motivation.}
+Reversible computation is an emerging computation paradigm encompassing
+computations that are not just deterministic when executed the \emph{forward}
+direction, but also in the \emph{backward} direction. While this may seem
+initially obscure, reversible computation forms the basis for quantum
+computing, and has seen applications in a number of different areas such as
+low-power computing~\cite{Landauer61}, robotics~\cite{SchultzLE18}, discrete
+event simulation~\cite{Schordan15}, and the simultaneous construction of
+parser/pretty printer pairs~\cite{RendelO10}. Like classical computing, it has
+its own automata~\cite{Bennett73}, circuit model~\cite{DeVos10}, machine
+architectures, programming
+languages~\cite{JamesS12,JamesS14,YokoyamaG07,Schultz18,JacobsenKT18}, semantic
+metalanguages~\cite{Giles,KaarsgaardAG17,KaarsgaardG18}, and so on.
 
-\todo{Formalization motivation? Maybe?}
+\todo{Niccol\`{o}: Formalization motivation? Maybe?}
 
-\todo{Explain Carette et al. Say that we are extending it to be actually useful (but don't say it like that).}
+\Pio{} is a family of reversible combinator calculi comprising structural
+isomorphisms and combinators corresponding to those found in dagger-traced
+$\omega$-continuous rig categories~\cite{Karvonen19}. Though superficially
+simple, \Pio{} is expressive enough as a metalanguage to give semantics to the
+typed reversible functional programming language Theseus~\cite{JamesS14}.
 
-\todo{Delay vs. Maybe monad.}
+In \cite{CaretteS16}, the authors studied the equivalences of isomorphisms in
+the finitary fragment of \Pio{} (i.e., without recursive types and iteration
+via the \trace{} combinator), and showed that these equivalences could be
+adequately described by another combinator calculus of equivalences of
+isomorphisms, in sum yielding a two-level calculus of isomorphisms and
+equivalences of isomorphisms. In this paper, we build on this work to produce a
+(fully formalized) two-level calculus for full \Pio{} (supporting both recursive
+types and iteration) via the \Delay{} monad, using insights gained from the
+study its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well as
+of join inverse categories in which reversible
+iteration may be modelled~\cite{KaarsgaardAG17}. \todo{Niccol\`{o}, do you have anything to add here?}
+
+\todo{Niccol\`{o}: \Delay vs. Maybe monad.}
 
@@
 & \infer{\unfold : \mu\,A \lr \sub \, A \, (\mu \,A)}{} \\
-\multicolumn{2}{|c|}{\infer{\trace \,f: B \lr C}{f : A \oplus B \lr A \oplus C}} \\
+\multicolumn{2}{|c|}{\infer{\trace \,f: A \lr B}{f : A \oplus C \lr B \oplus C}} \\
 \hline
@@
 \begin{align*}
-& \traceD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
-& \traceD \,f = \copairD{\iterD \,\mapL f}{\now} \diamond \mapR f
+& \traceD : (A + C \to \Delay (B + C)) \to A \to \Delay B \\
+& \traceD \,f = \copairD{\now}{\iterD \,\mapR f}\diamond \mapL f
 \end{align*}
@@
 \begin{align*}
-& \traceHD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
-& \traceHD \,f = \iterD (\Delay (\inl + \id) \comp f) \comp \inr
+& \traceHD : (A + C \to \Delay (B + C)) \to A \to \Delay B \\
+& \traceHD \,f = \iterD (\Delay (\id + \inr) \comp f) \comp \inl
 \end{align*}

### 2019-04-25 mpc19/paper.tex
Added rough abstract.

@@
 %
-\documentclass[runningheads]{llncs}
+\documentclass[runningheads,a4paper]{llncs}
 %
@@
 \begin{abstract}
-\todo{Write it!}
+Reversible computation studies computations which exhibit both forward and
+backward determinism. Among others, it has been studied for half a century for
+its applications in low-power computing, and forms the basis for quantum
+computing.
+
+Though certified program equivalence is useful for a number of applications
+(e.g., certified compilation and optimization), little work on this topic has
+been carried out for reversible programming languages. As a notable exception,
+Carette and Sabry have studied the equivalences of the finitary fragment of \Pio, a reversible combinator calculus, yielding a two-level calculus of type isomorphisms and equivalences between them.
+
+In this paper, we extend the two-level calculus of finitary \Pio{} to one for
+full \Pio{} (i.e., with both recursive types and iteration by means of a
+\trace{} combinator) using the \Delay{} monad, which can be regarded as a
+``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
+partiality. This yields a calculus of iterative (and possibly nonterminating)
+reversible programs acting on user-defined dynamic data structures together
+with a calculus of certified program equivalences between these programs.
+
+\todo{Niccol\`o, anything to add? Feel free to edit.}
 

### 2019-04-25 mpc19/paper.tex
working on dagger trace

--- 2019-04-24 mpc19/paper.tex
@@
 \newcommand{\traceHD}{\mathsf{traceH}_\D}
+\newcommand{\daggerD}{\mathsf{dagger}_\D}
+\newcommand{\Orb}[3]{\mathsf{Orb}\,#1\,#2\,#3}
+\newcommand{\done}{\mathsf{done}}
+\renewcommand{\next}{\mathsf{next}}
+\newcommand{\reverseOrbit}{\mathsf{reverseOrb}}
 
@@
 returning a value $b$, then $g\,b$ terminates returning $a$, and vice
-versa. In $\Dapprox$, our definition of partial isomorphisms is equivalent to
+versa. Given a partial isomorphism $f$, we denote its partial inverse
+by $\daggerD\,f$.
+
+In $\Dapprox$, our definition of partial isomorphisms is equivalent to
 the standard categorical one~\cite{Kastl79} (see also \cite{CockettL02}),
@@
 
-An Elgot monad is a monad $T$ whose Kleilsi category supports
-unguarded Elgot iteration. More precisely\footnote{Here we give the
-  definition of Elgot monad on $\Set$, but the definition of Elgot
-  monad makes sense in any category with binary coproducts.}, a monad
-$\T$ is Elgot if there exists an operation
+An Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
+whose Kleilsi category supports unguarded uniform iteration. More
+precisely\footnote{Here we give the definition of Elgot monad on
+  $\Set$, but the definition of Elgot monad makes sense in any
+  category with binary coproducts.}, a monad $\T$ is Elgot if there
+exists an operation
 \[
-\iter_\T : (A \to \T \,(A + B)) \to A \to \T\,B
+\iter_\T : (A \to \T \,(B + A)) \to A \to \T\,B
 \]
@@
 \[
-\iter_\T\,f \equiv \copair{\iter_\T \,f}{\eta_\T} \diamond_\T f
+\iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
 \]
@@
 \begin{array}{ll}
-\multicolumn{2}{l}{\iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B} \\
-\iterD' \,f \,(\now \,(\inl\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
-\iterD' \,f \,(\now \,(\inr\, b)) &= \now \,b \\
+\multicolumn{2}{l}{\iterD' : (A \to \Delay (B + A)) \to \Delay (B + A) \to \Delay B} \\
+\iterD' \,f \,(\now \,(\inl\, b)) &= \now \,b \\
+\iterD' \,f \,(\now \,(\inr\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
 \iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
  \\
-\multicolumn{2}{l}{\iterD : (A \to \Delay (A + B)) \to A \to \Delay B} \\
+\multicolumn{2}{l}{\iterD : (A \to \Delay (B + A)) \to A \to \Delay B} \\
 \multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
@@
 of $\later$ to the total computation time. This addition is necessary
-achieving the productivity of the corecursively defined function
-$\iterD'$. In fact, by changing the first line of its specification to
-$\iterD' \,f \,(\now \,(\inl\, a)) = \iterD' \,f \,(f a)$
-and taking $f = \inlD$, we would have that $\iterD' \,f \,(\now
-\,(\inl\, a))$ unfolds indefinitely without producing any output. In
+for ensuring the productivity of the corecursively defined function
+$\iterD'$. In fact, by changing the second line of its specification to
+$\iterD' \,f \,(\now \,(\inr\, a)) = \iterD' \,f \,(f a)$
+and taking $f = \inrD$, we would have that $\iterD' \,f \,(\now
+\,(\inr\, a))$ unfolds indefinitely without producing any output. In
 Agda, such a definition would be rightfully rejected by the termination
@@
 \begin{array}{ll}
-\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (A + B)) \to
-  \bind\, \copairD{\iterD \,f}{\now} \approx \iterD' \,f} \\
-\unfolding' \,f \,(\now \,(\inl \,a)) & = \laterR_{\approx} \,
+\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (B + A)) \to
+  \bind\, \copairD {\now}{\iterD \,f} \approx \iterD' \,f} \\
+\unfolding' \,f \,(\now \,(\inl \,b)) & = \now_\approx \\
+\unfolding' \,f \,(\now \,(\inr \,a)) & = \laterR_{\approx} \,
   \refl_\approx \\
-\unfolding' \,f \,(\now \,(\inr \,b)) & = \now_\approx \\
 \unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
@@
  \\
-\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (A + B)) \to \copairD{\iterD
-  \,f}{\now} \diamond f \approx \iterD \,f}\\
+\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (B + A)) \to
+  \copairD {\now}{\iterD \,f}\diamond f \approx \iterD \,f}\\
 \multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
@@
 $\mapR f = f \comp \inr : B \to C$, so that $f = [\mapL f , \mapR f]$.
+Graphically:
+\ctikzfig{fL_fR}
+
 The trace operator in $\Dapprox$ is defined as follows:
 \begin{align*}
-& \traceD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
-& \traceD \,f = \copairD{\iterD \,\mapL f}{\now} \diamond \mapR f
+& \traceD : (A + C \to \Delay (B + C)) \to A \to \Delay B \\
+& \traceD \,f = \copairD{\now}{\iterD \,\mapR f}\diamond \mapL f
 \end{align*}
 The operation $\traceD$ is compatible with weak bisimilarity. 
+Pictorially:
 
-\ctikzfig{fL_fR}
 \ctikzfig{trace}
+
+Intuitively, the function $\mapL f$ initialises the loop. It either
+diverges, so that the trace of $f$ diverges as well, or it terminates. It either terminates with
+an element $b : B$, in which case the loop ends immediately returning
+$b$, or it converges to a value $c : C$, and in this case we proceed
+by  invoking the iteration of $\mapR f$ on $c$.
 
@@
 operator, as shown by Hasegawa  \cite{Hasegawa97}. His construction,
-instantiated to our setting, looks like this:
+instantiated to our setting, looks as follows:
 \begin{align*}
-& \traceHD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
-& \traceHD \,f = \iterD (\Delay (\inl + \id) \comp f) \comp \inr
+& \traceHD : (A + C \to \Delay (B + C)) \to A \to \Delay B \\
+& \traceHD \,f = \iterD (\Delay (\id + \inr) \comp f) \comp \inl
 \end{align*}
 
+It is not difficoult to prove that the two possible ways of defining a
+trace operator from Elgot iteration are equivalent, in the sense that
+$\traceD\,f \approx \traceHD\,f$ for all $f : A + C \to \Delay (B +
+C)$.
+
+The trace axioms follow from the Elgot iteration axioms.
+
 \subsection{Dagger Trace}\label{sec:daggertrace}
+
+We now show that $\traceD$ is a $\dagger$-trace operator,
+i.e. if $f$ is a partial isomorphism, then
+$\traceD\,f$ is also a partial isomorphism with partial inverse $\traceD\,(\daggerD\,f)$.
+
+This is proved by introducing the notion of \emph{orbit} of an element
+$x : A + C$ wrt. a function $f : A + C \to \Delay (B + C)$. The orbit of
+$x$ consists of the terms of type $B + C$ that are obtained in a
+finite number of steps from repeated applications of the function $f$
+on $x$. Formally, a term $y$ belongs to the orbit of $f$ wrt. $x$ if
+the type $\Orb f x y $ is inhabited, with the latter type
+inductively defined as: 
+\[
+\infer{\done\,p : \Orb f x y}{p : f \,x \dn y}
+\qquad 
+\infer{\next\,p \,q : \Orb f x y}{p : f \,x \dn \inr\,c & q : \Orb f
+  {(\inr\,c)} y}
+\]
+
+The notion of orbit can be used to state when 
+the iteration of a function $f : A \to \Delay (B + A)$ on a input
+$a : A$ terminates with value $b : B$.
+\[
+\iterD\,f\,a \dn b \leftrightarrow \Orb {\copairD{\inlD}{f}} {(\inr\,a)} {(\inl\,b)}
+\]
+We refer the intersted reader to our Agda formalization for a complete
+proof of this logical equivalence.
+Similarly, the orbit can be used to state when 
+the trace of a function $f : A + C \to \Delay (B + C)$ on a input
+$a : A$ terminates with value $b : B$.
+\begin{equation}\label{eq:traceorbit}
+\traceD\,f\,a \dn b \leftrightarrow \Orb f{(\inl\,a)} {(\inl\,b)}
+\end{equation}
+
+Showing that $\traceD$ is a $\dagger$-trace operator requires the
+construction of an inhabitant of $\traceD\,f\,a \dn b \leftrightarrow
+\traceD\,(\daggerD\,f)\,b \dn a$. Thanks to the logical equivalence in
+(\ref{eq:traceorbit}), this is equivalent to prove the following
+statement instead: 
+\[
+\Orb f {(\inl\,a)} {(\inl\,b)} \leftrightarrow \Orb {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}
+\]
+We give a proof of the left-to-right direction, the other implication
+is proved in an analogous way.  Notice that a term
+$p : \Orb f {(\inl\,a)} {(\inl\,b)}$ can be seen as a finite sequence of
+elements of $C$, precisely the intermediate values produced by
+$\traceD\,f\,a$ before converging to $b$. The orbit of $b$ wrt. the
+partial inverse of $f$ can therefore be computed by reversing the
+sequence of elements present in $p$. The construction of the reverse
+of an orbit is very similar to the way the reverse of a list is typically
+defined in a functional programming language like Haskell.
+
+\[
+\def\arraystretch{1.2}
+\begin{array}{ll}
+\multicolumn{2}{l}{\reverseOrbit' : (i : \{a : A\} \{b : B\} \to f \,a
+\dn b \to \daggerD\,f  \,b \dn a) \to } \\
+\multicolumn{2}{l}{\qquad \{a : A\} \{b : B\} \{c : C\} \to } \\
+\multicolumn{2}{l}{\qquad \Orb f {(\inr\,c)}
+ {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inr\,c)} {(\inl\,a)} \to } \\
+\multicolumn{2}{l}{\qquad \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
+\reverseOrbit' \,i\, (\done \,p) \,p'&= \next\, (i \,p)\,p' \\
+\reverseOrbit' \,i\,(\next\,p\,q) \,p'&= \reverseOrbit' \,i\,q \,(\next \,(i\,p)\,p')
+\end{array}
+\]
+
+\[
+\def\arraystretch{1.2}
+\begin{array}{ll}
+\multicolumn{2}{l}{\reverseOrbit : (i : \{a : A\} \{b : B\} \to f \,a
+  \dn b \to \daggerD\,f  \,b \dn a) \to} \\
+\multicolumn{2}{l}{\qquad \{a : A\} \{b : B\} \to} \\
+\multicolumn{2}{l}{\qquad \Orb f {(\inl\,a)}
+  {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
+\reverseOrbit \,i\, (\done \,p) &= \done \, (i \,p) \\
+\reverseOrbit \,i\,(\next\,p\,q) &= \reverseOrbit' \,i\,q \,(\done \,(i\,p))
+\end{array}
+\]
 

### 2019-04-25 mpc19/paper.tex

@@
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
+Reversible computation studies computations which exhibit both forward and
+backward determinism. Among others, it has been studied for half a century for
+its applications in low-power computing, and forms the basis for quantum
+computing.
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+Though certified program equivalence is useful for a number of applications
+(e.g., certified compilation and optimization), little work on this topic has
+been carried out for reversible programming languages. As a notable exception,
+Carette and Sabry have studied the equivalences of the finitary fragment of \Pio, a reversible combinator calculus, yielding a two-level calculus of type isomorphisms and equivalences between them.
+
+In this paper, we extend the two-level calculus of finitary \Pio{} to one for
+full \Pio{} (i.e., with both recursive types and iteration by means of a
+\trace{} combinator) using the \Delay{} monad, which can be regarded as a
+``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
+partiality. This yields a calculus of iterative (and possibly nonterminating)
+reversible programs acting on user-defined dynamic data structures together
+with a calculus of certified program equivalences between these programs.
+
+\todo{Niccol\`o, anything to add? Feel free to edit.}
+
+\keywords{reversible computation \and iteration \and delay monad}
 \end{abstract}

### 2019-04-26 mpc19/paper.tex
more intro

@@
 \newcommand{\Delay}{\ensuremath{\mathsf{Delay}\,}}
+\newcommand{\Maybe}{\mathsf{Maybe}}
 \newcommand{\now}{\mathsf{now}}
@@
 
-\todo{Niccol\`{o}: Formalization motivation? Maybe?}
-
 \Pio{} is a family of reversible combinator calculi comprising structural
@@
 types and iteration) via the \Delay{} monad, using insights gained from the
-study its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well as
+study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well as
 of join inverse categories in which reversible
-iteration may be modelled~\cite{KaarsgaardAG17}. \todo{Niccol\`{o}, do you have anything to add here?}
+iteration may be modelled~\cite{KaarsgaardAG17}.
 
-\todo{Niccol\`{o}: \Delay vs. Maybe monad.}
+The coinductive delay datatype was introduced by Capretta
+\cite{Capretta05} as a way of defining possibly non-terminating
+functions in Martin-L\"of type theory. It has found a number of
+applications, ranging from operational semantics of functional
+languages \cite{Danielsson12} to formalization of domain theory in
+type theory \cite{BentonKV09} and normalization by evaluation
+\cite{AbelC14}. Here we employ it for giving denotational semantics to
+\Pio. In particular, we show how to endow the delay datatype with a
+dagger trace combinator, whose construction factors through the
+specification of a uniform iteration operator
+\cite{GoncharovSRJ18,GoncharovMR16}.
+
+Throughout the paper, we reason constructively in Martin-L\"of type
+theory. Classically, the delay monad is isomorphic to the maybe monad
+$\Maybe\,X \equiv X + 1$,
+and thus just a complication of something that can be expressed much
+simpler. But constructively they are very different. In particular,
+it is impossible to define a well-behaved trace combinator for the
+maybe monad without assuming classical principles such as the law of
+excluded middle.
+
+We have fully formalized the development of the paper in the
+dependently typed programming language Agda~\cite{Agda}. The code is
+available online at \\ \url{https://github.com/niccoloveltri/pi0-agda}. The
+formalization uses Agda 2.6.0.
 

### 2019-04-26 mpc19/paper.tex
Added missing } in bibtex library.
Updated title, minor edits.

@@
 %
-\title{Type-theoretical Semantics of \Pio\ via the Delay Monad \thanks{Supported by organization x.}}
+\title{En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Supported by organization x.}}
 %
-%\titlerunning{Abbreviated paper title}
+\titlerunning{Unguarded iteration for reversible computation in the Delay monad}
 % If the paper title is too long for the running head, you can set
@@
 ``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
-partiality. This yields a calculus of iterative (and possibly nonterminating)
+partiality. This yields a calculus of iterative (and possibly non-terminating)
 reversible programs acting on user-defined dynamic data structures together
 with a calculus of certified program equivalences between these programs.
-
-\todo{Niccol\`o, anything to add? Feel free to edit.}
 
@@
 \begin{itemize}
-  \item Abstract.
-  \item Introduction.
   \item Make pictures! 
@@
 its own automata~\cite{Bennett73}, circuit model~\cite{DeVos10}, machine
-architectures, programming
+architectures~\cite{ThomsenAG11}, programming
 languages~\cite{JamesS12,JamesS14,YokoyamaG07,Schultz18,JacobsenKT18}, semantic
@@
 types and iteration) via the \Delay{} monad, using insights gained from the
-study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well as
-of join inverse categories in which reversible
-iteration may be modelled~\cite{KaarsgaardAG17}.
+study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well
+as of join inverse categories in which reversible iteration may be
+modelled~\cite{KaarsgaardAG17}.
 
@@
 and thus just a complication of something that can be expressed much
-simpler. But constructively they are very different. In particular,
+simpler. Constructively, however, they are very different. In particular,
 it is impossible to define a well-behaved trace combinator for the
@@
 dependently typed programming language Agda~\cite{Agda}. The code is
-available online at \\ \url{https://github.com/niccoloveltri/pi0-agda}. The
+available online at \\ \texttt{\url{https://github.com/niccoloveltri/pi0-agda}}. The
 formalization uses Agda 2.6.0.

### 2019-04-26 mpc19/paper.tex
2programs

--- 2019-04-26 mpc19/paper.tex
@@
 \newcommand{\I}{\mathsf{I}}
+\newcommand{\LR}{\iff}
+\renewcommand{\dagger}{\mathsf{dagger}}
 
@@
 \begin{abstract}
-Reversible computation studies computations which exhibit both forward and
-backward determinism. Among others, it has been studied for half a century for
-its applications in low-power computing, and forms the basis for quantum
-computing.
+  Reversible computation studies computations which exhibit both
+  forward and backward determinism. Among others, it has been studied
+  for half a century for its applications in low-power computing, and
+  forms the basis for quantum computing.
 
-Though certified program equivalence is useful for a number of applications
-(e.g., certified compilation and optimization), little work on this topic has
-been carried out for reversible programming languages. As a notable exception,
-Carette and Sabry have studied the equivalences of the finitary fragment of \Pio, a reversible combinator calculus, yielding a two-level calculus of type isomorphisms and equivalences between them.
+  Though certified program equivalence is useful for a number of
+  applications (e.g., certified compilation and optimization), little
+  work on this topic has been carried out for reversible programming
+  languages. As a notable exception, Carette and Sabry have studied
+  the equivalences of the finitary fragment of \Pio, a reversible
+  combinator calculus, yielding a two-level calculus of type
+  isomorphisms and equivalences between them.
 
-In this paper, we extend the two-level calculus of finitary \Pio{} to one for
-full \Pio{} (i.e., with both recursive types and iteration by means of a
-\trace{} combinator) using the \Delay{} monad, which can be regarded as a
-``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
-partiality. This yields a calculus of iterative (and possibly nonterminating)
-reversible programs acting on user-defined dynamic data structures together
-with a calculus of certified program equivalences between these programs.
-
-\todo{Niccol\`o, anything to add? Feel free to edit.}
+  In this paper, we extend the two-level calculus of finitary \Pio{}
+  to one for full \Pio{} (i.e., with both recursive types and
+  iteration by means of a \trace{} combinator) using the \Delay{}
+  monad, which can be regarded as a ``computability-aware'' analogue
+  of the usual $\mathsf{Maybe}$ monad for partiality. This yields a
+  calculus of iterative (and possibly nonterminating) reversible
+  programs acting on user-defined dynamic data structures together
+  with a calculus of certified program equivalences between these
+  programs.
 
@@
 A ::= \Z \, | \,A \oplus A \, | \,\I \,| \,A \otimes A \,| \,X \,|
-\,\mu X.\,A 
+\,\mu X.A 
 \]
 
-%% \begin{figure}
-%% \[
-%% \arraycolsep=8pt\def\arraystretch{2.8}
-%% \begin{array}{|ccc|}
-%% \hline 
-%% \infer{\Z : \Ty_n}{} 
-%% & \infer{\I : \Ty_n}{}
-%% & \infer{\Var\,i : \Ty_n}{i \in [0..n)}  \\
-%% \infer{A \oplus B : \Ty_n}{A : \Ty_n & B : \Ty_n}
-%% & \infer{A \otimes B : \Ty_n}{A : \Ty_n & B : \Ty_n}
-%% & \infer{\mu\,A : \Ty_n}{A : \Ty_{n+1}} \\
-%% \hline
-%% \end{array}
-%% \]
-%% \caption{Types of \Pio.}
-%% \label{fig:types}
-%% \end{figure}
+\begin{figure}
+\[
+\arraycolsep=8pt\def\arraystretch{2.8}
+\begin{array}{ccc}
+%\hline 
+\infer{\Z : \Ty\,n}{} 
+& \infer{\I : \Ty\,n}{}
+& \infer{\Var\,i : \Ty\,n}{i \in [0..n)}  \\
+\infer{A \oplus B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
+& \infer{A \otimes B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
+& \infer{\mu A : \Ty\,n}{A : \Ty\,(n+1)} 
+%\\ \hline
+\end{array}
+\]
+\caption{Types of \Pio, as formalized in Agda}
+\label{fig:types}
+\end{figure}
 
@@
 \[
-\arraycolsep=8pt\def\arraystretch{2.8}
-\begin{array}{|cc|}
-\hline
-\infer{\id : A \lr A}{}
-& \infer{g \bullet f : A \lr C}{g : B \lr C & f : A \lr B} \\
+\arraycolsep=8pt\def\arraystretch{1.8}
+\begin{array}{cc}%{|cc|}
+%\hline 
+\id : A \lr A
+& \infer{g \bullet f : A \lr C}{g : B \lr C & f : A \lr B} \\ [.25cm]
 \infer{f \oplus g : A \oplus B \lr C \oplus D}{f : A \lr C & g : B \lr D} 
 & \infer{f \otimes g : A \otimes B \lr C \otimes D}{f : A \lr C & g : B \lr D} \\
-\infer{\lambda_{\oplus} : \Z \oplus A \lr A}{}
-& \infer{\lambda_{\oplus}^{-1} : A \lr \Z \oplus A}{} \\
-\infer{\lambda_{\otimes} : \I \otimes A \lr A}{}
-& \infer{\lambda_{\otimes}^{-1} : A \lr \I \otimes A}{} \\
-\infer{\alpha_{\oplus} : (A \oplus B) \oplus C \lr A \oplus (B \oplus C)}{}
-& \infer{\alpha_{\oplus}^{-1} : A \oplus (B \oplus C) \lr (A \oplus B) \oplus C}{} \\
-\infer{\alpha_{\otimes} : (A \otimes B) \otimes C \lr A \otimes (B \otimes C)}{}
-& \infer{\alpha_{\otimes}^{-1} : A \otimes (B \otimes C) \lr (A \otimes B) \otimes C}{} \\
-\infer{\sigma_{\oplus} : A \oplus B \lr B \oplus A}{}
-& \infer{\sigma_{\otimes} : A \otimes B \lr B \otimes A}{} \\
-\infer{\kappa : \Z \otimes A \lr \Z}{} 
-& \infer{\delta : (A \oplus B) \otimes C \lr (A \otimes C) \oplus (B \otimes C)}{} \\
-\infer{\kappa^{-1} : \Z \lr \Z \otimes A}{}
-& \infer{\delta^{-1} : (A \otimes C) \oplus (B \otimes C)  \lr (A \oplus B) \otimes C}{} \\
-\infer{\fold : \sub \, A \, (\mu \,A) \lr \mu \,A}{} 
-& \infer{\unfold : \mu\,A \lr \sub \, A \, (\mu \,A)}{} \\
-\multicolumn{2}{|c|}{\infer{\trace \,f: A \lr B}{f : A \oplus C \lr B \oplus C}} \\
-\hline
+\lambda_{\oplus} : \Z \oplus A \lr A
+& \lambda_{\oplus}^{-1} : A \lr \Z \oplus A\\
+\lambda_{\otimes} : \I \otimes A \lr A
+& \lambda_{\otimes}^{-1} : A \lr \I \otimes A\\
+\alpha_{\oplus} : (A \oplus B) \oplus C \lr A \oplus (B \oplus C)
+& \alpha_{\oplus}^{-1} : A \oplus (B \oplus C) \lr (A \oplus B) \oplus C \\
+\alpha_{\otimes} : (A \otimes B) \otimes C \lr A \otimes (B \otimes C)
+& \alpha_{\otimes}^{-1} : A \otimes (B \otimes C) \lr (A \otimes B) \otimes C \\
+\sigma_{\oplus} : A \oplus B \lr B \oplus A
+& \sigma_{\otimes} : A \otimes B \lr B \otimes A \\
+\kappa : \Z \otimes A \lr \Z 
+& \delta : (A \oplus B) \otimes C \lr (A \otimes C) \oplus (B \otimes C) \\
+\kappa^{-1} : \Z \lr \Z \otimes A
+& \delta^{-1} : (A \otimes C) \oplus (B \otimes C)  \lr (A \oplus B) \otimes C \\
+\fold : A[\mu X.A/X] \lr \mu X.A
+& \unfold : \mu X.A \lr A[\mu X.A/X] \\[.25cm]
+\multicolumn{2}{c}{\infer{\trace \,f: A \lr B}{f : A \oplus C \lr B \oplus C}} \\
+%\hline
 \end{array}
@@
 \label{fig:programs}
+\end{figure}
+
+%% \infer{\lambda_{\oplus} : \Z \oplus A \lr A}{}
+%% & \infer{\lambda_{\oplus}^{-1} : A \lr \Z \oplus A}{} \\
+%% \infer{\lambda_{\otimes} : \I \otimes A \lr A}{}
+%% & \infer{\lambda_{\otimes}^{-1} : A \lr \I \otimes A}{} \\
+%% \infer{\alpha_{\oplus} : (A \oplus B) \oplus C \lr A \oplus (B \oplus C)}{}
+%% & \infer{\alpha_{\oplus}^{-1} : A \oplus (B \oplus C) \lr (A \oplus B) \oplus C}{} \\
+%% \infer{\alpha_{\otimes} : (A \otimes B) \otimes C \lr A \otimes (B \otimes C)}{}
+%% & \infer{\alpha_{\otimes}^{-1} : A \otimes (B \otimes C) \lr (A \otimes B) \otimes C}{} \\
+%% \infer{\sigma_{\oplus} : A \oplus B \lr B \oplus A}{}
+%% & \infer{\sigma_{\otimes} : A \otimes B \lr B \otimes A}{} \\
+%% \infer{\kappa : \Z \otimes A \lr \Z}{} 
+%% & \infer{\delta : (A \oplus B) \otimes C \lr (A \otimes C) \oplus (B \otimes C)}{} \\
+%% \infer{\kappa^{-1} : \Z \lr \Z \otimes A}{}
+%% & \infer{\delta^{-1} : (A \otimes C) \oplus (B \otimes C)  \lr (A \oplus B) \otimes C}{} \\
+%% \infer{\fold : \sub \, A \, (\mu \,A) \lr \mu \,A}{} 
+%% & \infer{\unfold : \mu\,A \lr \sub \, A \, (\mu \,A)}{} \\
+%% \multicolumn{2}{c}{\infer{\trace \,f: A \lr B}{f : A \oplus C \lr B \oplus C}} \\
+
+\subsection{Program Equivalences}
+
+\begin{figure}
+\[
+\arraycolsep=8pt\def\arraystretch{1.8}
+\begin{array}{cc} %{|cc|} \hline 
+\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
+  \id) \bullet g)} \\ 
+\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
+  \oplus \id))} \\  
+\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
+  \,(g \bullet (\id \oplus f))} \\
+\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
+  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
+\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
+  \bullet \rho_\oplus 
+&
+\mathsf{yanking} : \trace \sigma_\oplus \LR \id \\
+\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet (\dagger \,f)
+  \bullet g \bullet (\dagger \, g) \LR g \bullet (\dagger \,g)
+  \bullet f \bullet (\dagger \, f)} \\
+\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
+&
+\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
+\\[.25cm]
+\multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
+  \bullet \trace\,f \LR \trace \,f}{p : f \bullet (\dagger\,f) \bullet
+  f \LR f}}
+%\\ \hline
+\end{array}
+\]
+\caption{Selection of program equivalences of \Pio}
+\label{fig:programequivs}
 \end{figure}
@@
 
-We now show that $\traceD$ is a $\dagger$-trace operator,
+We now show that $\traceD$ is a dagger trace operator,
 i.e. if $f$ is a partial isomorphism, then
@@
 
-Showing that $\traceD$ is a $\dagger$-trace operator requires the
+Showing that $\traceD$ is a dagger trace operator requires the
 construction of an inhabitant of $\traceD\,f\,a \dn b \leftrightarrow
@@
 is proved in an analogous way.  Notice that a term
-$p : \Orb f {(\inl\,a)} {(\inl\,b)}$ can be seen as a finite sequence of
-elements of $C$, precisely the intermediate values produced by
+$p : \Orb f {(\inl\,a)} {(\inl\,b)}$ can be seen as a finite sequence
+of elements of $C$, precisely the intermediate values produced by
 $\traceD\,f\,a$ before converging to $b$. The orbit of $b$ wrt. the
@@
 sequence of elements present in $p$. The construction of the reverse
-of an orbit is very similar to the way the reverse of a list is typically
-defined in a functional programming language like Haskell.
-
+of an orbit is very similar to the way the reverse of a list is
+typically defined in a functional programming language like Haskell.
+We first consider an intermediate value $c : C$ and we assume to have
+already reversed the initial section of the orbit between $\inl\,a$
+and $\inr\,c$, that is a term $p' : \Orb  {(\daggerD\,f)} {(\inr\,c)}
+{(\inl\,a)}$. 
 \[
 \def\arraystretch{1.2}
-\begin{array}{ll}
-\multicolumn{2}{l}{\reverseOrbit' : (i : \{a : A\} \{b : B\} \to f \,a
+\begin{array}{lll}
+\multicolumn{3}{l}{\reverseOrbit' : (i : \{a : A\} \{b : B\} \to f \,a
 \dn b \to \daggerD\,f  \,b \dn a) \to } \\
-\multicolumn{2}{l}{\qquad \{a : A\} \{b : B\} \{c : C\} \to } \\
-\multicolumn{2}{l}{\qquad \Orb f {(\inr\,c)}
+\multicolumn{3}{l}{\qquad \{a : A\} \{b : B\} \{c : C\} \to } \\
+\multicolumn{3}{l}{\qquad \Orb f {(\inr\,c)}
  {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inr\,c)} {(\inl\,a)} \to } \\
-\multicolumn{2}{l}{\qquad \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
-\reverseOrbit' \,i\, (\done \,p) \,p'&= \next\, (i \,p)\,p' \\
-\reverseOrbit' \,i\,(\next\,p\,q) \,p'&= \reverseOrbit' \,i\,q \,(\next \,(i\,p)\,p')
+\multicolumn{3}{l}{\qquad \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
+\reverseOrbit' \,i\, (\done \,p) & p'&= \next\, (i \,p)\,p' \\
+\reverseOrbit' \,i\,(\next\,p\,q) & p'&= \reverseOrbit' \,i\,q \,(\next \,(i\,p)\,p')
 \end{array}
 \]
+The proof of $\reverseOrbit'$ proceeds by structural induction on the final
+segment of the orbit between $\inr\,c$ and $\inl\,b$ that still needs
+to be reversed, which is the argument of type
+$\Orb f {(\inr\,c)} {(\inl\,b)}$. There are two possibilities.
+\begin{itemize}
+\item We have $p : f\,(\inr\,c) \dn
+\inl\,b$, in which case $i\,p : \daggerD\,f\,(\inl\,b) \dn
+\inl\,c$. Then we return $\next\,(i\,p)\,p'$.
+\item There exists another value $c' : C$ such that $p :
+  f\,(\inr\,c) \dn \inr\,c'$ and $q : \Orb f {(\inr\,c')}
+  {(\inl\,b)}$. Then we recursively invoke the function
+  $\reverseOrbit' \,i$ on arguments $q$ and $\next\,(i\,p)\,p' : \Orb
+  {(\daggerD \,f)} {(\inr\,c')} {(\inl\,b)}$.
+\end{itemize}
 
+The reverse of an orbit is derivable using the auxiliary function $\reverseOrbit'$.
 \[
@@
 \]
+The proof of $\reverseOrbit$ proceeds bystructural induction on the
+orbit of type
+$\Orb f {(\inl\,a)} {(\inl\,b)}$. There are two possibilities.
+\begin{itemize}
+\item We have $p : f\,(\inl\,a) \dn
+\inl\,b$, in which case $i\,p : \daggerD\,f\,(\inl\,b) \dn
+\inl\,a$. Then we return $\done\,(i\,p)$.
+\item There exists a value $c : C$ such that $p :
+  f\,(\inl\,a) \dn \inr\,c$ and $q : \Orb f {(\inr\,c)}
+  {(\inl\,b)}$. We conclude by invoking the function
+  $\reverseOrbit' \,i$ on arguments $q$ and $\done\,(i\,p) : \Orb
+  {(\daggerD \,f)} {(\inr\,c)} {(\inl\,a)}$.
+\end{itemize}
 

### 2019-04-26 mpc19/paper.tex

@@
 \begin{abstract}
-  Reversible computation studies computations which exhibit both
-  forward and backward determinism. Among others, it has been studied
-  for half a century for its applications in low-power computing, and
-  forms the basis for quantum computing.
+Reversible computation studies computations which exhibit both forward and
+backward determinism. Among others, it has been studied for half a century for
+its applications in low-power computing, and forms the basis for quantum
+computing.
 
-  Though certified program equivalence is useful for a number of
-  applications (e.g., certified compilation and optimization), little
-  work on this topic has been carried out for reversible programming
-  languages. As a notable exception, Carette and Sabry have studied
-  the equivalences of the finitary fragment of \Pio, a reversible
-  combinator calculus, yielding a two-level calculus of type
-  isomorphisms and equivalences between them.
+Though certified program equivalence is useful for a number of applications
+(e.g., certified compilation and optimization), little work on this topic has
+been carried out for reversible programming languages. As a notable exception,
+Carette and Sabry have studied the equivalences of the finitary fragment of \Pio, a reversible combinator calculus, yielding a two-level calculus of type isomorphisms and equivalences between them.
 
-  In this paper, we extend the two-level calculus of finitary \Pio{}
-  to one for full \Pio{} (i.e., with both recursive types and
-  iteration by means of a \trace{} combinator) using the \Delay{}
-  monad, which can be regarded as a ``computability-aware'' analogue
-  of the usual $\mathsf{Maybe}$ monad for partiality. This yields a
-  calculus of iterative (and possibly nonterminating) reversible
-  programs acting on user-defined dynamic data structures together
-  with a calculus of certified program equivalences between these
-  programs.
+In this paper, we extend the two-level calculus of finitary \Pio{} to one for
+full \Pio{} (i.e., with both recursive types and iteration by means of a
+\trace{} combinator) using the \Delay{} monad, which can be regarded as a
+``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
+partiality. This yields a calculus of iterative (and possibly non-terminating)
+reversible programs acting on user-defined dynamic data structures together
+with a calculus of certified program equivalences between these programs.
 

### 2019-04-26 mpc19/paper.tex
Types and terms of pi0

@@
 
+The types of \Pio\ are given by the grammar:
 \[
@@
 \]
+where $X$ ranges over a set of variables. In Agda, we use de Bruijn
+indexes to deal with type variables, so the grammar above is formally
+realized by the rules in Figure \ref{fig:types}. The type $\Ty\,n$
+represents \Pio\ types containing at most $n$ free
+variables. Variables themselves are encoded as elements of
+$\mathsf{Fin}\,n$, the type of natural numbers strictly smaller then
+$n$. Notice that the $\mu$ binds a variable, which, for $A : \Ty\,(n +
+1)$, we consider to be $n + 1$.
 
@@
 & \infer{\I : \Ty\,n}{}
-& \infer{\Var\,i : \Ty\,n}{i \in [0..n)}  \\
+& \infer{\Var\,i : \Ty\,n}{i : \mathsf{Fin}\,n}  \\
 \infer{A \oplus B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
@@
 
-\subsection{Programs}
+\subsection{Terms}
 
+The terms of \Pio\ are inductively generated by the rules in Figure
+\ref{fig:programs}. They include the identity programs $\id$ and
+sequential composition of programs $\bullet$. $(\Z,\oplus)$ is a
+symmetric monoidal structure, with terms $\lambda_\oplus$,
+$\alpha_\oplus$ and $\sigma_\otimes$ as structural
+morphisms. Similarly for $(\I,\otimes)$. Moreover $\otimes$
+distributes over $\Z$ and $\oplus$, as evidenced by  $\kappa$
+and $\delta$. Elements of $\mu X.A$ are built using the term
+constructor $\fold$ and destructed with $\unfold$. Finally, we find the
+$\trace$ combinator.
+
+Every \Pio\ program is reversible. The inverse of program is given by
+the function $\dagger : (A \lr B) \to (B \lr A)$ recursively defined below:
 \begin{figure}
 \[
-\arraycolsep=8pt\def\arraystretch{1.8}
+\arraycolsep=7pt\def\arraystretch{1.8}
 \begin{array}{cc}%{|cc|}
@@
 \]
-\caption{Programs of \Pio}
+\caption{Terms of \Pio}
 \label{fig:programs}
 \end{figure}
+
+\[
+\begin{array}{llcll}
+%\multicolumn{2}{l}{\dagger : \{A B : \Ty \,n\} \to A \lr B \to B \lr A} \\
+\dagger \,\id &= \id & \quad &
+\dagger\, (g \bullet f) & = \dagger\, f \bullet \dagger\, g \\
+\dagger\, (f \oplus g) & = \dagger\, f \oplus \dagger\, g & \quad &
+\dagger\, (f \otimes g) & = \dagger\, f \otimes \dagger\, g \\
+\dagger\, \lambda_\oplus ^{-1} & = \lambda_\oplus & \quad &
+\dagger\, \lambda_\oplus & = \lambda_\oplus^{-1} \\
+\dagger\, \sigma_\oplus & = \sigma_\oplus & \quad &
+\dagger\, \alpha_\oplus & = \alpha_\oplus^{-1} \\
+\dagger\, \alpha_\oplus^{-1} & = \alpha_\oplus & \quad &
+\dagger\, \lambda_\otimes & = \lambda_\otimes^{-1} \\
+\dagger\, \lambda_\otimes ^{-1} & = \lambda_\otimes & \quad &
+\dagger\, \sigma_\otimes & = \sigma_\otimes \\
+\dagger\, \alpha_\otimes & = \alpha_\otimes^{-1} & \quad &
+\dagger\, \alpha_\otimes^{-1} & = \alpha_\otimes \\
+\dagger\, \kappa & = \kappa^{-1} & \quad &
+\dagger\, \kappa^{-1} & = \kappa \\
+\dagger\, \delta & = \delta^{-1} & \quad &
+\dagger\, \delta^{-1} & = \delta \\
+\dagger\, \fold & = \unfold & \quad &
+\dagger\, \unfold & = \fold \\
+\dagger\, (\trace \,f) & = \trace \,(\dagger\, f)
+\end{array}
+\]
 
@@
 
-\subsection{Program Equivalences}
+\subsection{Term Equivalences}
 
@@
 \mathsf{yanking} : \trace \sigma_\oplus \LR \id \\
-\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet (\dagger \,f)
-  \bullet g \bullet (\dagger \, g) \LR g \bullet (\dagger \,g)
-  \bullet f \bullet (\dagger \, f)} \\
+\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet \dagger \,f
+  \bullet g \bullet \dagger \, g \LR g \bullet \dagger \,g
+  \bullet f \bullet \dagger \, f} \\
 \mathsf{foldIso} : \fold \bullet \unfold \LR \id 
@@
 \multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
-  \bullet \trace\,f \LR \trace \,f}{p : f \bullet (\dagger\,f) \bullet
+  \bullet \trace\,f \LR \trace \,f}{p : f \bullet \dagger\,f \bullet
   f \LR f}}
@@
 \]
-\caption{Selection of program equivalences of \Pio}
+\caption{Selection of term equivalences of \Pio}
 \label{fig:programequivs}
@@
 
-We now show that $\traceD$ is a dagger trace operator,
+We now move to show that $\traceD$ is a dagger trace operator,
 i.e. if $f$ is a partial isomorphism, then
@@
 
-\section{Interpretation}\label{sec:interpretation}
+\section{Soundness}\label{sec:interpretation}
 

### 2019-04-29 mpc19/paper.tex
worked on section 2

@@
 
+In this section, we present the syntax of \Pio. The 1-structure of
+\Pio, i.e. its types and terms, has originaly been introduced by James
+and Sabry \cite{JamesS12}. In particular, we include the presence of
+recursive types and a primitive trace combinator.  Following Carette
+and Sabry's formalization of the finitary fragment of \Pio, we
+consider a collection of equivalences between terms. Our list of
+axioms notably differs from theirs in that we do not require each term to be
+a total isomorphism, we ask only for the existence of a partial
+inverse.
+
+Formally, the collection of types of \Pio\ constitutes the free
+dagger-traced rig inverse category with recursive types. \todo{Robin: Does
+  this make any sense? Also, anything to add here? Like some facts
+  about \Pio, like Turing completeness.}
+
 \subsection{Types}
@@
 
+It is also necessary to define substitutions. In Agda, given
+types $A : \Ty\,(n+1)$ and $B : \Ty\,n$, we construct
+$\sub\,A\,B : \Ty\,n$ to represent the substituted type $A[B/X]$,
+where $X$ corresponds to the $(n+1)$-th variable in context.
+
 \subsection{Terms}
@@
 morphisms. Similarly for $(\I,\otimes)$. Moreover $\otimes$
-distributes over $\Z$ and $\oplus$, as evidenced by  $\kappa$
+distributes over $\Z$ and $\oplus$ from the right, as evidenced by  $\kappa$
 and $\delta$. Elements of $\mu X.A$ are built using the term
@@
 
-Every \Pio\ program is reversible. The inverse of program is given by
-the function $\dagger : (A \lr B) \to (B \lr A)$ recursively defined below:
 \begin{figure}
@@
 
+\begin{figure}
+\[
+\arraycolsep=8pt\def\arraystretch{1.8}
+\begin{array}{cc} %{|cc|} \hline 
+\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
+  \id) \bullet g)} \\ 
+\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
+  \oplus \id))} \\  
+\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
+  \,(g \bullet (\id \oplus f))} \\
+\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
+  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
+\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
+  \bullet \rho_\oplus 
+&
+\mathsf{yanking} : \trace \sigma_\oplus \LR \id
+ \\
+\multicolumn{2}{c}{\mathsf{tracePIso} : \trace\,f \bullet \trace\,(\dagger\,f)
+  \bullet \trace\,f \LR \trace \,f} \\
+%%  \\[.25cm]
+%% \multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
+%%   \bullet \trace\,f \LR \trace \,f}{p : f \bullet \dagger\,f \bullet
+%%   f \LR f}} \\
+\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
+&
+\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
+\\
+\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet \dagger \,f
+  \bullet g \bullet \dagger \, g \LR g \bullet \dagger \,g
+  \bullet f \bullet \dagger \, f}
+%\\ \hline
+\end{array}
+\]
+\caption{Selection of term equivalences of \Pio}
+\label{fig:programequivs}
+\end{figure}
+
+Every \Pio\ program is reversible. The (partial) inverse of a program is given by
+the function $\dagger : (A \lr B) \to (B \lr A)$, recursively defined
+as follows:
+
 \[
@@
 
+The $\dagger$ operation is involutive. Notice that this property holds
+up to propositional equality. This is proved by induction on the term
+$f$.
+\[
+\mathsf{daggerInvol} : (f : A \lr B) \to \dagger \,(\dagger\,f) \equiv f
+\]
+
+The right unitor for $\oplus$ is given by $\rho_\oplus = \lambda_\oplus \bullet
+\sigma_\oplus : A \oplus \Z \lr A,$ and $\rho_\otimes$ is defined similarly.
+Analogously, we can derive the left distributors $\kappa' : A \otimes \Z
+\lr A$ and $\delta' : A \otimes (B \oplus C) \lr (A \otimes B) \oplus
+(A \otimes C)$.
+
 %% \infer{\lambda_{\oplus} : \Z \oplus A \lr A}{}
@@
 
-\begin{figure}
+A selection of term equivalences of \Pio\ is given in Figure
+\ref{fig:programequivs}. We only include the equivalences that either
+differ or have not previously considered by Carette and Sabry in their
+formalization of the finite fragment of \Pio \cite{CaretteS16}. In
+particular, we leave out the long list of Laplaza's coherence axioms
+expressing that the types of \Pio\ form a rig category \cite{Laplaza}.
+We also omit the equivalences stating that $\lambda_\oplus^{-1}$ is
+the total inverse of $\lambda_\oplus$, similarly for the other
+structural morphisms.
+
+The list of term equivalences in Figure \ref{fig:programequivs}
+contains the trace axioms, displaying that the types of \Pio\ form a
+traced monoidal category wrt. the additive monoidal structure
+$(\Z,\oplus)$ \cite{Karvonen19}. Next we ask for
+$\trace\,(\dagger\,f)$ to be the partial inverse of
+$\trace\,f$. Remember that we have defined $\dagger\,(\trace\,f)$ to
+be $\trace\,(\dagger \,f)$, so the axiom $\mathsf{tracePIso}$
+is evidence that the trace combinator of \Pio\ is a dagger trace.
+Afterwards we have two equivalences stating that $\unfold$ is the
+total inverse of $\fold$. Finally,... 
+\todo{Robin: a nice description of the last equivalence in Figure 3?
+  Perhaps a citation too?}
+
+It is possible to show that every term $f$ has $\dagger\,f$ as its
+partial inverse, therefore proving that the types of \Pio\ possess an
+inverse category structure \todo{Am I right here?}. The proof proceeds by induction on $f$.
 \[
-\arraycolsep=8pt\def\arraystretch{1.8}
-\begin{array}{cc} %{|cc|} \hline 
-\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
-  \id) \bullet g)} \\ 
-\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
-  \oplus \id))} \\  
-\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
-  \,(g \bullet (\id \oplus f))} \\
-\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
-  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
-\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
-  \bullet \rho_\oplus 
-&
-\mathsf{yanking} : \trace \sigma_\oplus \LR \id \\
-\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet \dagger \,f
-  \bullet g \bullet \dagger \, g \LR g \bullet \dagger \,g
-  \bullet f \bullet \dagger \, f} \\
-\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
-&
-\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
-\\[.25cm]
-\multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
-  \bullet \trace\,f \LR \trace \,f}{p : f \bullet \dagger\,f \bullet
-  f \LR f}}
-%\\ \hline
-\end{array}
+\mathsf{existsPIso} : (f : A \lr B) \to f \bullet \dagger\,f
+\bullet\,f \LR f
 \]
-\caption{Selection of term equivalences of \Pio}
-\label{fig:programequivs}
-\end{figure}
 

### 2019-04-29 mpc19/paper.tex
Minor corrections.

@@
 \newcommand{\comp}{\circ}
-\newcommand{\U}{\mathcal{U}}
+% \newcommand{\U}{\mathcal{U}}
 \newcommand{\copair}[2]{[#1,#2]}
@@
 
-Formally, the collection of types of \Pio\ constitutes the free
-dagger-traced rig inverse category with recursive types. \todo{Robin: Does
-  this make any sense? Also, anything to add here? Like some facts
-  about \Pio, like Turing completeness.}
+Formally, the collection of types of \Pio\ correspond to those naturally
+interpreted in dagger traced $\omega$-continuous rig categories (see
+\cite{Karvonen19}).
 
@@
 Afterwards we have two equivalences stating that $\unfold$ is the
-total inverse of $\fold$. Finally,... 
-\todo{Robin: a nice description of the last equivalence in Figure 3?
-  Perhaps a citation too?}
+total inverse of $\fold$.
 
-It is possible to show that every term $f$ has $\dagger\,f$ as its
-partial inverse, therefore proving that the types of \Pio\ possess an
-inverse category structure \todo{Am I right here?}. The proof proceeds by induction on $f$.
+It is possible to show that every term $f$ has $\dagger\,f$ as its partial
+inverse, and since the final equivalence of Figure~\ref{fig:programequivs} is
+equivalent to stating that partial inverses are unique~\cite{Kastl79}, it
+follows that the types and terms of \Pio\ form an inverse category. The proof
+proceeds by induction on $f$.
 \[

### 2019-04-29 mpc19/paper.tex
soundness
Added diagram for Hasegawa trace.

@@
 \begin{itemize}
-  \item Make pictures! 
-  \begin{itemize}
-    \item Hasegawa trace from iterator.
-  \end{itemize}
   \item Conclusion
@@
 It is well-known that a trace operator is obtainable from an iteration
-operator, as shown by Hasegawa  \cite{Hasegawa97}. His construction,
+operator, as shown by Hasegawa~\cite{Hasegawa97}. His construction,
 instantiated to our setting, looks as follows:
@@
 \end{align*}
+or graphically
+\ctikzfig{trace_hasegawa}
 
-It is not difficoult to prove that the two possible ways of defining a
+It is not difficult to prove that the two possible ways of defining a
 trace operator from Elgot iteration are equivalent, in the sense that
@@
 \]
-We refer the intersted reader to our Agda formalization for a complete
+We refer the interested reader to our Agda formalization for a complete
 proof of this logical equivalence.

### 2019-04-29 mpc19/paper.tex
interpretation of terms

--- 2019-04-29 mpc19/paper.tex
@@
 \usepackage{fancyvrb}
+\usepackage{stmaryrd}
 
@@
 
+% Semantics
+\newcommand{\semTy}[1]{{\llbracket #1 \rrbracket}_\mathsf{Ty}}
+\newcommand{\semMu}[1]{{\llbracket #1 \rrbracket}_\mu}
+\newcommand{\semTm}[1]{{\llbracket #1 \rrbracket}_\lr}
+
 \newcommand{\todo}[1]{\hl{TODO: #1}}
@@
 \end{itemize}
-Niccolò: 
-\begin{itemize}
-  \item Fix the syntax to be closer to ordinary presentations (but do describe 
-  how it is realized in Agda).
-  \item Finish the section about trace and iteration.
-  \item Remind the reader about double lines of coinductive definitions.
-  \item Swap types in Elgot iterator.
-\end{itemize}
-Both:
-\begin{itemize}
-  \item \textbf{Important:} Come up with a punnier title! (Something about 
-  delays and daggers).
-\end{itemize}
+
 % section todo (end)
@@
 
+In this section, we provide some details on the interpretation of the
+syntax of \Pio, presented in Section \ref{sec:syntax}, into the
+category $\Inv\Dapprox$. Types of \Pio\ are modelled as objects of
+$\Inv\Dapprox$, which are types of the metatheory. In Agda, the
+interpretation of types $\semTy -$ takes in input a \Pio\ type
+$A : \Ty\,n$ and an environment $\rho : \mathsf{Fin}\,n \to \Set$
+giving semantics to each variable in context. The interpretation is
+mutually inductively defined together with the operation $\semMu -$
+giving semantics to the $\mu$ type former.
+\[
+\begin{array}{lll}
+\semTy \Z & \rho & = 0 \\
+\semTy {A \oplus B} & \rho & = \semTy A \,\rho+ \semTy B  \, \rho\\
+\semTy \I & \rho & = 1 \\
+\semTy {A \otimes B} & \rho & = \semTy A \,\rho \times \semTy B \,\rho\\
+\semTy {\Var\,i} & \rho & = \rho\, i \\
+\semTy {\mu A} & \rho & = \semMu A \,\rho
+\end{array}
+\qquad \qquad
+\infer{\mathsf{semFold} \,x : \semMu A\,\rho}{x : \semTy A \, \rho}
+\]
+
+Terms of \Pio\ are modelled as morphism of $\Inv\Dapprox$,
+i.e. partial isomorphisms. Formally, we define an operation:
+\[
+\semTm - : (A \lr B) \to 
+\]
+
 \section{Conclusions}

### 2019-04-29 mpc19/paper.tex
Added section on semantcis

@@
 \newcommand{\semMu}[1]{{\llbracket #1 \rrbracket}_\mu}
-\newcommand{\semTm}[1]{{\llbracket #1 \rrbracket}_\lr}
+\newcommand{\semTm}[1]{{\llbracket #1 \rrbracket}_{\lr}}
+\newcommand{\semEq}[1]{{\llbracket #1 \rrbracket}_{\LR}}
 
@@
 
+Summing up, in this section we have showed that the $\traceD$ operator
+can be restricted to act on partial isomorphisms. That is, the
+following type is inhabited:
+\[
+\mathsf{semTrace} : A + C\simeq B + C \to A \simeq C
+\]
+
 \section{Soundness}\label{sec:interpretation}
@@
 \[
-\semTm - : (A \lr B) \to 
-\]
+\semTm - : (A \lr B) \to \semTy A \, \rho \simeq \semTy B \, \rho
+\]
+We refer the interested reader to our Agda formalization for a
+complete definition of the interpretation of terms. We only emphasize
+the semantics of the trace operator: $\semTm {\trace\,f} =
+\mathsf{semTrace}\,\semTm f$.
+
+Term equivalences of \Pio\ are modelled as morphism equalities in
+$\Inv\Dapprox$, i.e. proofs of weak bisimilarity between two
+morpshisms. Formally, we define an operation:
+\[
+\semEq - : (f \LR g) \to \semTm f \approx \semTm g
+\]
+Again we refer the interested reader to our Agda formalization for a
+complete definition of the interpretation of term equivalences.
 

### 2019-04-29 mpc19/paper.tex
fixed weak bisim

@@
 \newcommand{\semEq}[1]{{\llbracket #1 \rrbracket}_{\LR}}
+\newcommand{\semTrace}{\mathsf{trace}_{\D\simeq}}
 
@@
 computation. Double rule lines refer to a coinductive constructor,
-which can be applied an infinite amount of times.
-The non-terminating computation $\never$
-is corecursively defined as $\never = \later \,\never$.
+which can be employed an infinite number of times in the construction
+of a term of type $\Delay A$.  E.g., the non-terminating computation
+$\never$ is corecursively defined as $\never = \later \,\never$.
 
@@
 Two computations in $\Delay A$ are weakly bisimilar if they differ by
-a finite number of applications of the constructor $\later$. This
+a finite number of applications of the constructor
+$\later$. Alternatively, we can say that two computations $x$ and $y$
+are weakly bisimilar if, whenever $x$ terminates returning a value
+$a$, then $y$ also terminates returning $a$, and vice versa. This
 informal statement can be formalized in several different but
-logically equivalent ways \cite{ChapmanUV19}. 
-A possible defintion is as follows:
-\begin{equation}\label{eq:weak1}
-x \approx y = (a : A) \to x \dn a \leftrightarrow y \dn a
-\end{equation}
-The definition says that $x$ and $y$ are weakly bisimilar if, whenever
-$x$ terminates returning a value $a$, then $y$ also terminates
-returning $a$, and vice versa.
-Alternatively, weak bisimilarity can be coinductively formulated as follows:
+logically equivalent ways \cite{ChapmanUV19}. Here we consider a
+coinductive formulation employed e.g. in \cite{Danielsson18}.
+%% A possible defintion is as follows:
+%% \begin{equation}\label{eq:weak1}
+%% x \approx y = (a : A) \to x \dn a \leftrightarrow y \dn a
+%% \end{equation}
+%% The definition says that $x$ and $y$ are weakly bisimilar if, whenever
+%% $x$ terminates returning a value $a$, then $y$ also terminates
+%% returning $a$, and vice versa.
+%% Alternatively, weak bisimilarity can be coinductively formulated as follows:
 \begin{equation}
@@
 \end{equation}
-Notice that the constructor $\later_{\approx}$ is
-coinductive. Formulation (\ref{eq:weak1}) and (\ref{eq:weak2}) are
-logically equivalent, meaning that two computations are weakly
-bisimilar as in (\ref{eq:weak1}) if and only if they are weakly
-bisimilar as in (\ref{eq:weak2}).
-In the rest of the paper we will employ
-both encodings of weak bisimilarity, in each case explicitly stating
-if we are working with definition (\ref{eq:weak1}) or definition (\ref{eq:weak2}).
-Weak bisimilarity is an equivalence
-relation and it is a congruence w.r.t. the $\later$ operation.
-For example, here is a proof that weak bisimilarity is reflexive,
-employing definition (\ref{eq:weak2}).
+Notice that the constructor $\later_{\approx}$ is coinductive. This
+allows us to prove $\never \approx \never$. 
+%% Formulation
+%% (\ref{eq:weak1}) and (\ref{eq:weak2}) are logically equivalent,
+%% meaning that two computations are weakly bisimilar as in
+%% (\ref{eq:weak1}) if and only if they are weakly bisimilar as in
+%% (\ref{eq:weak2}).  In the rest of the paper we will employ both
+%% encodings of weak bisimilarity, in each case explicitly stating if we
+%% are working with definition (\ref{eq:weak1}) or definition
+%% (\ref{eq:weak2}). 
+ Weak bisimilarity is an equivalence relation and it
+is a congruence w.r.t. the $\later$ operation.  For example, here is a
+proof that weak bisimilarity is reflexive.
 \begin{align*}
@@
 up to weak bisimilarity. Here we show the proof of the unfolding
-axiom. We employ the definition (\ref{eq:weak2}) of weak
-bisimilarity. The proof relies on an auxiliary proof $\unfolding'$. 
+axiom. The proof relies on an auxiliary proof $\unfolding'$. 
 \[
@@
 
+We conclude this section by remarking that the construction of a trace
+operator in the Kleisli category of the maybe monad is impossible
+without the assumption of additional classical principles. In fact,
+given a map $f : A + C \to B + C + 1$, let $xs$ be the possibly
+infinite sequence of elements of $B + C + 1$ produced by the iteration
+of $f$ on a given input $A + C$. In order to construct the trace of
+$f$, we need to decide whether $xs$ is a finite sequence terminating
+with an element of $B + 1$, or $xs$ is an infinite stream of elements
+of $C$. This decision requires the limited principle of omniscience,
+an instance of the law of excluded middle not provable in Martin-L\"of
+type theory.
+\[
+\mathsf{LPO} = (s : \mathbb{N} \to 2) \to ((n : \mathbb{N}) \times
+s\,n \equiv \mathsf{true}) + ((n : \mathbb{N}) \to s\,n \equiv \mathsf{false})
+\]
+
 \subsection{Dagger Trace}\label{sec:daggertrace}
@@
 \[
-\mathsf{semTrace} : A + C\simeq B + C \to A \simeq C
+\semTrace : A + C\simeq B + C \to A \simeq C
 \]
@@
 We refer the interested reader to our Agda formalization for a
-complete definition of the interpretation of terms. We only emphasize
+complete definition of the interpretation of terms. Here we limit
+ourselves to emphasize
 the semantics of the trace operator: $\semTm {\trace\,f} =
-\mathsf{semTrace}\,\semTm f$.
+\semTrace\,\semTm f$.
 

### 2019-04-29 mpc19/paper.tex
my acknowledgements

@@
 %
-\title{En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Supported by organization x.}}
+\title{En garde! Unguarded iteration for reversible computation in the
+  Delay monad\thanks{Niccol\`o Veltri was supported by a research
+    grant (13156) from VILLUM FONDEN.}}
 %
@@
 an instance of the law of excluded middle not provable in Martin-L\"of
-type theory.
+type theory:
 \[
@@
 \]
+where $2$ is the type of booleans, with $\mathsf{true}$ and
+$\mathsf{false}$ as only inhabitants.
 

### 2019-04-30 mpc19/paper.tex
conclusion on other monads of partiality

@@
 \newcommand{\iterD}{\mathsf{iter}_\D}
-\newcommand{\unfolding}{\mathsf{unfolding}}
+\newcommand{\unfolding}{\mathsf{fixpoint}_\D}
 \newcommand{\refl}{\mathsf{refl}}
@@
 informal statement can be formalized in several different but
-logically equivalent ways \cite{ChapmanUV19}. Here we consider a
-coinductive formulation employed e.g. in \cite{Danielsson18}.
+logically equivalent ways \cite{ChapmanUV19,VeltriPhD}. Here we
+consider a coinductive formulation employed e.g. in
+\cite{Danielsson18}.
 %% A possible defintion is as follows:
@@
 disputable semi-classical choice principles, we decided to quotient
-the homsets of $\D$ by (poitwise) weak bisimilarity instead of the
+the homsets of $\D$ by (pointwise) weak bisimilarity instead of the
 delay datatype.
@@
 $A + B + A \times B$. Nevertheless, it is possible to prove that 1 and
-$- \times -$ are partial terminal object and partial binary products
+$\times$ are partial terminal object and partial binary products
 respectively, in the sense of Cockett and Lack's restriction
@@
 \]
-satisfying a number of axioms, most notably the unfolding axiom:
-\[
+satisfying a number of axioms:
+\[\arraycolsep=8pt\def\arraystretch{1.8}
+\begin{array}{c}
+\mathsf{fixpoint} :
 \iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
+\\
+\mathsf{naturality} : g \diamond_\T \iter_\T\,f \equiv
+  \iter_\T\,(\copair{\T \inl \comp g} {\eta \comp \inr} \diamond_\T f)
+\\
+\mathsf{codiagonal} :
+\iter_\T \,(\iter_\T\,g) \equiv \iter_\T\,(\T \copair {\id} {\inr}
+  \comp g)
+\\[.25cm]
+\infer{\mathsf{uniformity}\,p : \iter_\T \,f \comp h \equiv \iter_\T
+  \,g}{p : f \comp h \equiv \T (\id + h) \comp g}
+\end{array}
 \]
@@
 composition in the Kleisli category of $\T$.
+The standard definition of uniform iteration operator includes the
+dinaturality axiom, which has recently been discovered to be derivable
+from the other laws \cite{EsikG16,GoncharovMR16}.
 
@@
 As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
-up to weak bisimilarity. Here we show the proof of the unfolding
+up to weak bisimilarity. Here we show the proof of the fixpoint
 axiom. The proof relies on an auxiliary proof $\unfolding'$. 
@@
 
+It is natural to wonder if our work can be ported to other monads of
+partiality in Martin-L\"of type theory. As already discussed in the
+end of Section \ref{sec:trace}, the maybe monad is not suitable for
+modelling a well-behaved trace combinator without the assumption of
+classical principles such as \textsf{LPO}. The partial map classifier
+\cite{UustaluV17-2,EscardoK17}
+$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where
+$\mathsf{Prop}$ is the type of propositions (types with at most one
+inhabitant), supports the existence of a uniform iteration operator
+and therefore a trace. Nevertheless, the specification of iteration is
+more complicated then the one presented in Section \ref{sec:elgot} for
+the delay monad, which is a simple corecursive definition. The Elgot
+monad structure of $\mathsf{PMC}$ follows from its Kleisli category
+being a join restriction category, so iteration is defined in terms of
+least upper bounds of certain chains of morphisms. The subcategory of
+partial isomorphisms of the Kleisli category of $\mathsf{PMC}$
+supports a dagger trace combinator, which can be proved following
+Kaarsgaard et al.'s general strategy \cite{KaarsgaardAG17}.  The exact
+same observations apply to the partiality monad in homotopy type
+theory \cite{AltenkirchDK17,ChapmanUV19}, to which the quotiented
+delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
+isomorphic to under the assumption of countable choice.
+
 %

### 2019-04-30 mpc19/paper.tex
Added a conclusion!

--- 2019-04-29 mpc19/paper.tex
@@
 \end{abstract}
-
-\section{TODO} % (fold)
-\label{sec:todo}
-Robin:
-\begin{itemize}
-  \item Conclusion
-  \begin{itemize}
-    \item Related work: Kleisli category of the Delay monad is a join
-    restriction category, inverse category of a join restriction category is a
-    join inverse category; already known to be dagger-traced under certain
-    conditions.
-    \item Future work: Semantics of time invertible programming. Extension of
-    the work on the Kleisli category of the Delay monad beyond $\Set$ to other
-    (more general) categories, e.g. topoi.
-  \end{itemize}
-\end{itemize}
-
-% section todo (end)
 
@@
 
+By abuse of notation, we use here $\cdot \times \cdot$ (respectively $\cdot +
+\cdot$) to refer to the product (respectively coproduct) in $\Dapprox$ even
+though it fails to be a product (respectively coproduct) in $\Inv\Dapprox$.
+However, both of these are symmetric monoidal products in $\Inv\Dapprox$, so
+their use as objects of $\Inv\Dapprox$ in the interpretation above is justified.
+
 Terms of \Pio\ are modelled as morphism of $\Inv\Dapprox$,
@@
 \section{Conclusions}
+In this paper, we have extended the work of Carette and Sabry~\cite{CaretteS16}
+to a (fully formalized) two-level calculus of \Pio{} programs and program
+equivalences. Key in this effort was the use of the Kleisli category of the
+$\Delay$ monad on $\Set$ under weak bisimilarity, which turned out to support
+iteration via a trace that preserves all partial isomorphisms, in this way
+giving semantics to the dagger trace of \Pio. Further, the work was formalized
+using Agda 2.6.0.
+
+The notion of computable partiality offered the $\Delay$ monad and its
+restriction category structure is well
+studied~\cite{VeltriPhD,UustaluV17,ChapmanUV19}. Indeed, though it could be
+established from previous work (see also \cite{KaarsgaardAG17}) that this
+category was traced, it was unclear whether the trace preserved partial
+isomorphisms -- a crucial point in its applicability to interpreting \Pio. The
+presented solution further avoids CPO-enrichment (as otherwise used in both
+\cite{UustaluV17} and \cite{KaarsgaardAG17}), and its issues in relation to
+formalization, in favor of a more direct definition of the trace from the Elgot
+iterator.
+
+Though the Kleisli category of the $\Delay$ monad on $\Set$ is well studied,
+comparatively less is known about this monad on other categories. It could be
+interesting to study under which conditions its iterator exist -- e.g., whether
+this is still the case when $\Set$ is replaced with an arbitrary topos. Another
+avenue concerns the study of \emph{time invertible} programming languages:
+Though not immediately clear in the current presentation, the trace on
+$\Inv\Dapprox$ is not just reversible (in the sense that it preserves partial
+isomorphisms) but in fact time invertible, in the sense that the number of
+computation steps needed to perform $\trace (\dagger f)$ is \emph{precisely}
+the same as what is needed to perform $\trace~f$ on any input. Since the
+$\Delay$ monad conveniently allows the counting of computation steps, we
+conjecture that this is an ideal setting in which to study such intentional
+semantics of reversible programming languages.
 

### 2019-04-30 mpc19/paper.tex

--- 2019-04-30 mpc19/paper.tex
@@
 \section{Conclusions}
-
-It is natural to wonder if our work can be ported to other monads of
-partiality in Martin-L\"of type theory. As already discussed in the
-end of Section \ref{sec:trace}, the maybe monad is not suitable for
-modelling a well-behaved trace combinator without the assumption of
-classical principles such as \textsf{LPO}. The partial map classifier
-\cite{UustaluV17-2,EscardoK17}
-$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where
-$\mathsf{Prop}$ is the type of propositions (types with at most one
-inhabitant), supports the existence of a uniform iteration operator
-and therefore a trace. Nevertheless, the specification of iteration is
-more complicated then the one presented in Section \ref{sec:elgot} for
-the delay monad, which is a simple corecursive definition. The Elgot
-monad structure of $\mathsf{PMC}$ follows from its Kleisli category
-being a join restriction category, so iteration is defined in terms of
-least upper bounds of certain chains of morphisms. The subcategory of
-partial isomorphisms of the Kleisli category of $\mathsf{PMC}$
-supports a dagger trace combinator, which can be proved following
-Kaarsgaard et al.'s general strategy \cite{KaarsgaardAG17}.  The exact
-same observations apply to the partiality monad in homotopy type
-theory \cite{AltenkirchDK17,ChapmanUV19}, to which the quotiented
-delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
+In this paper, we have extended the work of Carette and Sabry~\cite{CaretteS16}
+to a (fully formalized) two-level calculus of \Pio{} programs and program
+equivalences. Key in this effort was the use of the Kleisli category of the
+$\Delay$ monad on $\Set$ under weak bisimilarity, which turned out to support
+iteration via a trace that preserves all partial isomorphisms, in this way
+giving semantics to the dagger trace of \Pio. Further, the work was formalized
+using Agda 2.6.0.
+
+It is natural to wonder if our work can be ported to other monads of partiality
+in Martin-L\"of type theory. As already discussed in the end of Section
+\ref{sec:trace}, the $\Maybe$ monad is not suitable for modelling a
+well-behaved trace combinator without the assumption of classical principles
+such as \textsf{LPO}. The partial map classifier \cite{UustaluV17-2,EscardoK17}
+$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where $\mathsf{Prop}$ is
+the type of propositions (types with at most one inhabitant), supports the
+existence of a uniform iteration operator and therefore a trace. Nevertheless,
+the specification of iteration is more complicated then the one presented in
+Section \ref{sec:elgot} for the delay monad, which is a simple corecursive
+definition. The Elgot monad structure of $\mathsf{PMC}$ follows from its
+Kleisli category being a join restriction category, so iteration is defined in
+terms of least upper bounds of certain chains of morphisms. The subcategory of
+partial isomorphisms of the Kleisli category of $\mathsf{PMC}$ supports a
+dagger trace combinator, which can be proved following the general strategy in
+\cite{KaarsgaardAG17}. The exact same observations apply to the partiality
+monad in homotopy type theory \cite{AltenkirchDK17,ChapmanUV19}, to which the
+quotiented delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
 isomorphic to under the assumption of countable choice.
+
+Though the Kleisli category of the $\Delay$ monad on $\Set$ is well studied,
+comparatively less is known about this monad on other categories. It could be
+interesting to study under which conditions its iterator exist -- e.g., whether
+this is still the case when $\Set$ is replaced with an arbitrary topos. Another
+avenue concerns the study of \emph{time invertible} programming languages:
+Though not immediately clear in the current presentation, the trace on
+$\Inv\Dapprox$ is not just reversible (in the sense that it preserves partial
+isomorphisms) but in fact time invertible, in the sense that the number of
+computation steps needed to perform $\trace (\dagger f)$ is \emph{precisely}
+the same as what is needed to perform $\trace~f$ on any input. Since the
+$\Delay$ monad conveniently allows the counting of computation steps, we
+conjecture that this is an ideal setting in which to study such intentional
+semantics of reversible programming languages.
 

### 2019-04-30 mpc19/paper.tex
Homogenized our use of \Delay vs. delay, \Maybe vs. maybe, \trace vs. trace.

@@
 full \Pio{} (i.e., with both recursive types and iteration by means of a
-\trace{} combinator) using the \Delay{} monad, which can be regarded as a
-``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
+\trace{} combinator) using the delay monad, which can be regarded as a
+``computability-aware'' analogue of the usual maybe monad for
 partiality. This yields a calculus of iterative (and possibly non-terminating)
@@
 the finitary fragment of \Pio{} (i.e., without recursive types and iteration
-via the \trace{} combinator), and showed that these equivalences could be
+via the trace combinator), and showed that these equivalences could be
 adequately described by another combinator calculus of equivalences of
@@
 (fully formalized) two-level calculus for full \Pio{} (supporting both recursive
-types and iteration) via the \Delay{} monad, using insights gained from the
+types and iteration) via the delay monad, using insights gained from the
 study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well
@@
 \times B)$. Similarly, it has a unique costrength operation $\costr :
-\Delay A \times B \to \Delay (A \times B)$ definable using $\str$. Moreover, the delay datatype is a commutative monad.
+(\Delay A) \times B \to \Delay (A \times B)$ definable using $\str$. Moreover, the delay datatype is a commutative monad.
 
@@
 equivalences. Key in this effort was the use of the Kleisli category of the
-$\Delay$ monad on $\Set$ under weak bisimilarity, which turned out to support
+delay monad on $\Set$ under weak bisimilarity, which turned out to support
 iteration via a trace that preserves all partial isomorphisms, in this way
@@
 in Martin-L\"of type theory. As already discussed in the end of Section
-\ref{sec:trace}, the $\Maybe$ monad is not suitable for modelling a
+\ref{sec:trace}, the maybe monad is not suitable for modelling a
 well-behaved trace combinator without the assumption of classical principles
@@
 
-Though the Kleisli category of the $\Delay$ monad on $\Set$ is well studied,
+Though the Kleisli category of the delay monad on $\Set$ is well studied,
 comparatively less is known about this monad on other categories. It could be
@@
 the same as what is needed to perform $\trace~f$ on any input. Since the
-$\Delay$ monad conveniently allows the counting of computation steps, we
+delay monad conveniently allows the counting of computation steps, we
 conjecture that this is an ideal setting in which to study such intentional

### 2019-04-30 mpc19/paper.tex
minor

--- 2019-04-30 mpc19/paper.tex
@@
 isomorphisms) but in fact time invertible, in the sense that the number of
-computation steps needed to perform $\trace (\dagger f)$ is \emph{precisely}
-the same as what is needed to perform $\trace~f$ on any input. Since the
+computation steps needed to perform $\traceD \,(\daggerD\, f)$ is \emph{precisely}
+the same as what is needed to perform $\traceD\,f$ on any input. Since the
 $\Delay$ monad conveniently allows the counting of computation steps, we

### 2019-04-30 mpc19/paper.tex

--- 2019-04-30 mpc19/paper.tex
@@
 \]
-satisfying a number of axioms:
+satisfying the following axioms:
 \[\arraycolsep=8pt\def\arraystretch{1.8}
@@
 isomorphisms) but in fact time invertible, in the sense that the number of
-computation steps needed to perform $\trace (\dagger f)$ is \emph{precisely}
-the same as what is needed to perform $\trace~f$ on any input. Since the
+computation steps needed to perform $\traceD \,(\daggerD\, f)$ is \emph{precisely}
+the same as what is needed to perform $\traceD\,f$ on any input. Since the
 delay monad conveniently allows the counting of computation steps, we

### 2019-04-30 mpc19/paper.tex
Superposing

@@
 
+It is also necessary to define substitutions. In Agda, given
+types $A : \Ty\,(n+1)$ and $B : \Ty\,n$, we construct
+$\sub\,A\,B : \Ty\,n$ to represent the substituted type $A[B/X]$,
+where $X$ corresponds to the $(n+1)$-th variable in context.
+
 \begin{figure}
@@
 
-It is also necessary to define substitutions. In Agda, given
-types $A : \Ty\,(n+1)$ and $B : \Ty\,n$, we construct
-$\sub\,A\,B : \Ty\,n$ to represent the substituted type $A[B/X]$,
-where $X$ corresponds to the $(n+1)$-th variable in context.
-
 \subsection{Terms}
@@
 $\trace$ combinator.
+
+Every \Pio\ program is reversible. The (partial) inverse of a program is given by
+the function $\dagger : (A \lr B) \to (B \lr A)$, recursively defined
+as follows:
+
+\[
+\begin{array}{llcll}
+%\multicolumn{2}{l}{\dagger : \{A B : \Ty \,n\} \to A \lr B \to B \lr A} \\
+\dagger \,\id &= \id & \quad &
+\dagger\, (g \bullet f) & = \dagger\, f \bullet \dagger\, g \\
+\dagger\, (f \oplus g) & = \dagger\, f \oplus \dagger\, g & \quad &
+\dagger\, (f \otimes g) & = \dagger\, f \otimes \dagger\, g \\
+\dagger\, \lambda_\oplus ^{-1} & = \lambda_\oplus & \quad &
+\dagger\, \lambda_\oplus & = \lambda_\oplus^{-1} \\
+\dagger\, \sigma_\oplus & = \sigma_\oplus & \quad &
+\dagger\, \alpha_\oplus & = \alpha_\oplus^{-1} \\
+\dagger\, \alpha_\oplus^{-1} & = \alpha_\oplus & \quad &
+\dagger\, \lambda_\otimes & = \lambda_\otimes^{-1} \\
+\dagger\, \lambda_\otimes ^{-1} & = \lambda_\otimes & \quad &
+\dagger\, \sigma_\otimes & = \sigma_\otimes \\
+\dagger\, \alpha_\otimes & = \alpha_\otimes^{-1} & \quad &
+\dagger\, \alpha_\otimes^{-1} & = \alpha_\otimes \\
+\dagger\, \kappa & = \kappa^{-1} & \quad &
+\dagger\, \kappa^{-1} & = \kappa \\
+\dagger\, \delta & = \delta^{-1} & \quad &
+\dagger\, \delta^{-1} & = \delta \\
+\dagger\, \fold & = \unfold & \quad &
+\dagger\, \unfold & = \fold \\
+\dagger\, (\trace \,f) & = \trace \,(\dagger\, f)
+\end{array}
+\]
+
+The $\dagger$ operation is involutive. Notice that this property holds
+up to propositional equality. This is proved by induction on the term
+$f$.
+\[
+\mathsf{daggerInvol} : (f : A \lr B) \to \dagger \,(\dagger\,f) \equiv f
+\]
+
+The right unitor for $\oplus$ is given by $\rho_\oplus = \lambda_\oplus \bullet
+\sigma_\oplus : A \oplus \Z \lr A,$ and $\rho_\otimes$ is defined similarly.
+Analogously, we can derive the left distributors $\kappa' : A \otimes \Z
+\lr A$ and $\delta' : A \otimes (B \oplus C) \lr (A \otimes B) \oplus
+(A \otimes C)$.
 
@@
 
-\begin{figure}
-\[
-\arraycolsep=8pt\def\arraystretch{1.8}
-\begin{array}{cc} %{|cc|} \hline 
-\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
-  \id) \bullet g)} \\ 
-\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
-  \oplus \id))} \\  
-\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
-  \,(g \bullet (\id \oplus f))} \\
-\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
-  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
-\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
-  \bullet \rho_\oplus 
-&
-\mathsf{yanking} : \trace \sigma_\oplus \LR \id
- \\
-\multicolumn{2}{c}{\mathsf{tracePIso} : \trace\,f \bullet \trace\,(\dagger\,f)
-  \bullet \trace\,f \LR \trace \,f} \\
-%%  \\[.25cm]
-%% \multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
-%%   \bullet \trace\,f \LR \trace \,f}{p : f \bullet \dagger\,f \bullet
-%%   f \LR f}} \\
-\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
-&
-\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
-\\
-\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet \dagger \,f
-  \bullet g \bullet \dagger \, g \LR g \bullet \dagger \,g
-  \bullet f \bullet \dagger \, f}
-%\\ \hline
-\end{array}
-\]
-\caption{Selection of term equivalences of \Pio}
-\label{fig:programequivs}
-\end{figure}
-
-Every \Pio\ program is reversible. The (partial) inverse of a program is given by
-the function $\dagger : (A \lr B) \to (B \lr A)$, recursively defined
-as follows:
-
-\[
-\begin{array}{llcll}
-%\multicolumn{2}{l}{\dagger : \{A B : \Ty \,n\} \to A \lr B \to B \lr A} \\
-\dagger \,\id &= \id & \quad &
-\dagger\, (g \bullet f) & = \dagger\, f \bullet \dagger\, g \\
-\dagger\, (f \oplus g) & = \dagger\, f \oplus \dagger\, g & \quad &
-\dagger\, (f \otimes g) & = \dagger\, f \otimes \dagger\, g \\
-\dagger\, \lambda_\oplus ^{-1} & = \lambda_\oplus & \quad &
-\dagger\, \lambda_\oplus & = \lambda_\oplus^{-1} \\
-\dagger\, \sigma_\oplus & = \sigma_\oplus & \quad &
-\dagger\, \alpha_\oplus & = \alpha_\oplus^{-1} \\
-\dagger\, \alpha_\oplus^{-1} & = \alpha_\oplus & \quad &
-\dagger\, \lambda_\otimes & = \lambda_\otimes^{-1} \\
-\dagger\, \lambda_\otimes ^{-1} & = \lambda_\otimes & \quad &
-\dagger\, \sigma_\otimes & = \sigma_\otimes \\
-\dagger\, \alpha_\otimes & = \alpha_\otimes^{-1} & \quad &
-\dagger\, \alpha_\otimes^{-1} & = \alpha_\otimes \\
-\dagger\, \kappa & = \kappa^{-1} & \quad &
-\dagger\, \kappa^{-1} & = \kappa \\
-\dagger\, \delta & = \delta^{-1} & \quad &
-\dagger\, \delta^{-1} & = \delta \\
-\dagger\, \fold & = \unfold & \quad &
-\dagger\, \unfold & = \fold \\
-\dagger\, (\trace \,f) & = \trace \,(\dagger\, f)
-\end{array}
-\]
-
-The $\dagger$ operation is involutive. Notice that this property holds
-up to propositional equality. This is proved by induction on the term
-$f$.
-\[
-\mathsf{daggerInvol} : (f : A \lr B) \to \dagger \,(\dagger\,f) \equiv f
-\]
-
-The right unitor for $\oplus$ is given by $\rho_\oplus = \lambda_\oplus \bullet
-\sigma_\oplus : A \oplus \Z \lr A,$ and $\rho_\otimes$ is defined similarly.
-Analogously, we can derive the left distributors $\kappa' : A \otimes \Z
-\lr A$ and $\delta' : A \otimes (B \oplus C) \lr (A \otimes B) \oplus
-(A \otimes C)$.
-
 %% \infer{\lambda_{\oplus} : \Z \oplus A \lr A}{}
@@
 
+\begin{figure}
+\[
+\arraycolsep=8pt\def\arraystretch{1.8}
+\begin{array}{cc} %{|cc|} \hline 
+\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
+  \id) \bullet g)} \\ 
+\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
+  \oplus \id))} \\  
+\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
+  \,(g \bullet (\id \oplus f))} \\
+\multicolumn{2}{c}{\mathsf{superposing} : \trace \,(\alpha_\oplus^{-1}
+  \bullet (\id \oplus f) \bullet \alpha_\oplus) \LR \id \oplus \trace \,f} \\ 
+\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
+  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
+\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
+  \bullet \rho_\oplus 
+&
+\mathsf{yanking} : \trace\, \sigma_\oplus \LR \id
+ \\
+\multicolumn{2}{c}{\mathsf{tracePIso} : \trace\,f \bullet \trace\,(\dagger\,f)
+  \bullet \trace\,f \LR \trace \,f} \\
+%%  \\[.25cm]
+%% \multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
+%%   \bullet \trace\,f \LR \trace \,f}{p : f \bullet \dagger\,f \bullet
+%%   f \LR f}} \\
+\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
+&
+\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
+\\
+\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet \dagger \,f
+  \bullet g \bullet \dagger \, g \LR g \bullet \dagger \,g
+  \bullet f \bullet \dagger \, f}
+%\\ \hline
+\end{array}
+\]
+\caption{Selection of term equivalences of \Pio}
+\label{fig:programequivs}
+\end{figure}
+
 \subsection{Term Equivalences}
@@
 differ or have not previously considered by Carette and Sabry in their
-formalization of the finite fragment of \Pio \cite{CaretteS16}. In
+formalization of the finite fragment of \Pio~\cite{CaretteS16}. In
 particular, we leave out the long list of Laplaza's coherence axioms
@@
 traced monoidal category wrt. the additive monoidal structure
-$(\Z,\oplus)$ \cite{Karvonen19}. Next we ask for
+$(\Z,\oplus)$ \cite{JoyalSV96}. Next we ask for
 $\trace\,(\dagger\,f)$ to be the partial inverse of

### 2019-04-30 mpc19/paper.tex
more intro

@@
 \newcommand{\piso}{\mathsf{isPartialIso}}
-\newcommand{\pisoalt}{\mathsf{isPartialIso}_2}
+\newcommand{\pisoalt}{\mathsf{isPartialIsoCat}}
 \newcommand{\iterD}{\mathsf{iter}_\D}
@@
 
-The coinductive delay datatype was introduced by Capretta
-\cite{Capretta05} as a way of defining possibly non-terminating
-functions in Martin-L\"of type theory. It has found a number of
-applications, ranging from operational semantics of functional
-languages \cite{Danielsson12} to formalization of domain theory in
-type theory \cite{BentonKV09} and normalization by evaluation
-\cite{AbelC14}. Here we employ it for giving denotational semantics to
-\Pio. In particular, we show how to endow the delay datatype with a
-dagger trace combinator, whose construction factors through the
-specification of a uniform iteration operator
+The full \Pio\ calculus cannot be modelled in the same framework of
+\cite{CaretteS16}, since Martin-L\"of type theory is a total language
+which in particolar disallow the specification of a trace
+operator. Consequently, it is necessary to move to a setting
+supporting the existence of partial maps, and in type theory this can
+be done by using monads, by considering partiality as an effect. Our
+choice fell on the coinductive delay monad, introduced by Capretta
+\cite{Capretta05} as a way of representing general recursive functions
+in Martin-L\"of type theory. The delay datatype has been employed in a
+large number of applications, ranging from operational semantics of
+functional languages \cite{Danielsson12} to formalization of domain
+theory in type theory \cite{BentonKV09} and normalization by
+evaluation \cite{AbelC14}. Here it is used for giving denotational
+semantics to \Pio. In particular, we show how to endow the delay
+datatype with a trace combinator, whose construction factors
+through the specification of a uniform iteration operator
 \cite{GoncharovSRJ18,GoncharovMR16}.
@@
 it is impossible to define a well-behaved trace combinator for the
-maybe monad without assuming classical principles such as the law of
-excluded middle.
+maybe monad without assuming classical principles such as the limited
+principle of omniscience.
 
@@
 $\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
-\never : A \to \Delay 0$ as the terminal morphism. The
-binary product of $A$ and $B$ in $\Dapprox$ is given by
-$A + B + A \times B$. Nevertheless, it is possible to prove that 1 and
+\never : A \to \Delay 0$ as the terminal morphism. 
+%The binary product of $A$ and $B$ in $\Dapprox$ is given by $A + B + A \times B$. 
+Nevertheless, it is possible to prove that 1 and
 $\times$ are partial terminal object and partial binary products

### 2019-05-03 mpc19/paper.tex
Added figure for iterator.
small fixes

@@
 %
-\title{En garde! Unguarded iteration for reversible computation in the
-  Delay monad\thanks{Niccol\`o Veltri was supported by a research
+\title{En Garde! Unguarded Iteration for Reversible Computation in the
+  Delay Monad\thanks{Niccol\`o Veltri was supported by a research
     grant (13156) from VILLUM FONDEN.}}
@@
 full \Pio{} (i.e., with both recursive types and iteration by means of a
-\trace{} combinator) using the delay monad, which can be regarded as a
+trace combinator) using the delay monad, which can be regarded as a
 ``computability-aware'' analogue of the usual maybe monad for
@@
 \cite{CaretteS16}, since Martin-L\"of type theory is a total language
-which in particolar disallow the specification of a trace
-operator. Consequently, it is necessary to move to a setting
+which in particolar disallows the specification of a trace
+operator on types. Consequently, it is necessary to move to a setting
 supporting the existence of partial maps, and in type theory this can
@@
 projections are denoted $\fst$ and $\snd$. The empty type is 0 and the
-empty type is 1. We write $\Set$ for the category of
-types and functions between them.
+unit type is 1. We write $\Set$ for the category of
+types and functions between them. We also use $\Set$ to denote the
+universe of types.
 We define $A \leftrightarrow B = (A \to B) \times (B \to A)$.
@@
 $\mathsf{Fin}\,n$, the type of natural numbers strictly smaller then
-$n$. Notice that the $\mu$ binds a variable, which, for $A : \Ty\,(n +
+$n$. The type constructor $\mu$ binds a variable, which, for $A : \Ty\,(n +
 1)$, we consider to be $n + 1$.
@@
 particular, we leave out the long list of Laplaza's coherence axioms
-expressing that the types of \Pio\ form a rig category \cite{Laplaza}.
+expressing that types and terms of \Pio\ form a rig category \cite{Laplaza}.
 We also omit the equivalences stating that $\lambda_\oplus^{-1}$ is
@@
 partiality, which in particular allows the specification of a
-well-behaved trace operator, we introduce the notion of termination-sensitive
+well-behaved trace operator, we introduce the notion of (termination-sensitive)
 weak bisimilarity.
@@
 disputable semi-classical choice principles, we decided to quotient
-the homsets of $\D$ by (pointwise) weak bisimilarity instead of the
-delay datatype.
+the homsets of $\D$ by (pointwise) weak bisimilarity instead of
+the objects of $\D$.
 
@@
 binary coproduct of $A$ and $B$ in $\Dapprox$. Given $f : A \to
-\Delay C$ and $g : B \to \Delay C$, we define their copairing as
+\Delay C$ and $g : B \to \Delay C$, their copairing is
 $\copairD f g = \copair f g : A + B \to \Delay C$. The operation
@@
 interaction of partiality (in the sense of restriction categories) and limits
-in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is not
+in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is
+not the
 terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
@@
 The delay monad is an Elgot monad for which the axioms holds up to
-weak bisimilarity, not propositional equality. Its iteration operator
-relies on the definition of an auxiliary function $\iterD'$
-corecursively defined as follows:
+weak bisimilarity, not propositional equality. In other words, the
+category $\Dapprox$ can be endowed with a uniform iteration
+operator. The specification of the iteration operator relies on an
+auxiliary function $\iterD'$ corecursively defined as follows:
 \[ %\arraycolsep=8pt
@@
 up to weak bisimilarity. Here we show the proof of the fixpoint
-axiom. The proof relies on an auxiliary proof $\unfolding'$. 
+axiom, which in turns relies on an auxiliary proof $\unfolding'$. 
 \[
@@
 
-The trace operator in $\Dapprox$ is defined as follows:
+The trace operator in $\Dapprox$ is defined in terms of the iterator as follows:
 \begin{align*}
@@
 \end{align*}
-The operation $\traceD$ is compatible with weak bisimilarity. 
-Pictorially:
+The operation $\traceD$ is compatible with weak bisimilarity. Graphically, we express the iterator on $f$ as a wire looping back on the input, i.e., as
+
+\ctikzfig{iter}
+In this way, the definition of $\traceD$ may be expressed graphically as
 
 \ctikzfig{trace}
-
 Intuitively, the function $\mapL f$ initialises the loop. It either
@@
 infinite sequence of elements of $B + C + 1$ produced by the iteration
-of $f$ on a given input $A + C$. In order to construct the trace of
+of $f$ on a given input in $A$. In order to construct the trace of
 $f$, we need to decide whether $xs$ is a finite sequence terminating
@@
 \]
-We give a proof of the left-to-right direction, the other implication
-is proved in an analogous way.  Notice that a term
+We give a detailed proof of the left-to-right direction, the other implication
+is derived  in an analogous way.  Notice that a term
 $p : \Orb f {(\inl\,a)} {(\inl\,b)}$ can be seen as a finite sequence
@@
 \]
-The proof of $\reverseOrbit$ proceeds bystructural induction on the
+The proof of $\reverseOrbit$ proceeds by structural induction on the
 orbit of type
@@
 
-By abuse of notation, we use here $\cdot \times \cdot$ (respectively $\cdot +
-\cdot$) to refer to the product (respectively coproduct) in $\Dapprox$ even
-though it fails to be a product (respectively coproduct) in $\Inv\Dapprox$.
-However, both of these are symmetric monoidal products in $\Inv\Dapprox$, so
-their use as objects of $\Inv\Dapprox$ in the interpretation above is justified.
+By abuse of notation, we use here $\times$ (respectively $+$) to refer
+to the product (respectively coproduct) in $\Dapprox$ even though it
+fails to be a product (respectively coproduct) in $\Inv\Dapprox$.
+However, both of these are symmetric monoidal products in
+$\Inv\Dapprox$, so their use as objects of $\Inv\Dapprox$ in the
+interpretation above is justified.
 
@@
 quotiented delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
-isomorphic to under the assumption of countable choice.
+isomorphic under the assumption of countable choice.
 

### 2019-05-06 mpc19/paper.tex
overview + term interpretation

@@
 \newcommand{\semTm}[1]{{\llbracket #1 \rrbracket}_{\lr}}
-\newcommand{\semEq}[1]{{\llbracket #1 \rrbracket}_{\LR}}
+\newcommand{\semEq}[1]{{\llbracket #1 \rrbracket}_{\!\!\LR}}
 \newcommand{\semTrace}{\mathsf{trace}_{\D\simeq}}
@@
 \cite{CaretteS16}, since Martin-L\"of type theory is a total language
-which in particolar disallows the specification of a trace
+which in particular disallows the specification of a trace
 operator on types. Consequently, it is necessary to move to a setting
@@
 formalization uses Agda 2.6.0.
+
+\paragraph{Overview}
+In Section \ref{sec:syntax}, we present the syntax of \Pio\ as
+formalized in Agda, with particular emphasis on recursive types and
+the trace operator. In Section \ref{sec:delay}, we recall the
+definition of Capretta's delay datatype and weak bisimilarity. We
+discuss finite products and coproducts in the Kleisli category of the
+delay monad and we introduce the category of partial isomorphisms that
+serves as the denotational model of \Pio. In Section
+\ref{sec:elgot}, we build a complete Elgot monad structure on the delay
+datatype, that allows the encoding of a dagger trace operator in the
+category of partial isomorphisms. In Section \ref{sec:interpretation},
+we formally describe the interpretation of \Pio\ types, terms and
+terms equivalences. We conclude in Section \ref{sec:conclusions} with
+some final remarks and discussion on future work.
 
@@
 In this section, we present the syntax of \Pio. The 1-structure of
-\Pio, i.e. its types and terms, has originaly been introduced by James
+\Pio, i.e. its types and terms, has originally been introduced by James
 and Sabry \cite{JamesS12}. In particular, we include the presence of
@@
 
-\begin{figure}
+\begin{figure}[h]
 \[
@@
 
-\begin{figure}
+\begin{figure}[t]
 \[
@@
 We call $\Inv \Dapprox$ the subcategory of $\Dapprox$ consisting of
-(equivalence classes of) partial isomorphisms. Note that $\Inv\Dapprox$
-inherits neither partial products nor coproducts of $\Dapprox$, as the
-universal mapping property fails in both cases. However, it can be shown that
-in the category $\Inv\Dapprox$, $0$ is a zero object, $A + B$ is the
-disjointness tensor product of $A$ and $B$ (in the sense of Giles~\cite{Giles})
-with unit $0$, and $A \times B$ a monoidal product of $A$ and $B$ with unit $1$
-(though it is \emph{not} an inverse product in the sense of
-Giles~\cite{Giles}, as that would imply decidable equality on all objects).
+(equivalence classes of) partial isomorphisms. Note that
+$\Inv\Dapprox$ inherits neither partial products nor coproducts of
+$\Dapprox$, as the universal mapping property fails in both
+cases. However, it can be shown that in the category $\Inv\Dapprox$,
+$0$ is a zero object, $A + B$ is the disjointness tensor product of
+$A$ and $B$ (in the sense of Giles~\cite{Giles}) with unit $0$, and
+$A \times B$ a monoidal product of $A$ and $B$ with unit $1$ (though
+it is \emph{not} an inverse product in the sense of
+Giles~\cite{Giles}, as that would imply decidable equality on all
+objects).  In particular, we can derive the following operations,
+modelling the \Pio\ term constructors $\oplus$ and $\otimes$:
+\begin{align*}
+& \times_{\D \simeq} : A \simeq C \to B \simeq D \to A \times B
+  \simeq C \times D \\
+& +_{\D \simeq} : A \simeq C \to B \simeq D \to A + B
+  \simeq C + D
+\end{align*}
 
@@
 
-An Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
-whose Kleilsi category supports unguarded uniform iteration. More
-precisely\footnote{Here we give the definition of Elgot monad on
-  $\Set$, but the definition of Elgot monad makes sense in any
-  category with binary coproducts.}, a monad $\T$ is Elgot if there
+A complete Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
+whose Kleisli category supports unguarded uniform iteration. More
+precisely\footnote{Here we give the definition of complete Elgot monad on
+  $\Set$, but the definition of complete Elgot monad makes sense in any
+  category with finite coproducts.}, a monad $\T$ is Elgot if there
 exists an operation
@@
 
-The delay monad is an Elgot monad for which the axioms holds up to
+The delay monad is a complete Elgot monad for which the axioms holds up to
 weak bisimilarity, not propositional equality. In other words, the
@@
 The definition above can be given the following intuitive
-explaination. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
+explanation. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
 not terminate as well. If $f\,a$ terminates, there are two
@@
 $\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
-to $\inl\,a'$, in which case we repeat the precedure by replacing $a$
+to $\inl\,a'$, in which case we repeat the procedure by replacing $a$
 with $a'$. Notice that in the latter case we also add one occurrence
@@
 Terms of \Pio\ are modelled as morphism of $\Inv\Dapprox$,
-i.e. partial isomorphisms. Formally, we define an operation:
-\[
-\semTm - : (A \lr B) \to \semTy A \, \rho \simeq \semTy B \, \rho
-\]
-We refer the interested reader to our Agda formalization for a
-complete definition of the interpretation of terms. Here we limit
-ourselves to emphasize
-the semantics of the trace operator: $\semTm {\trace\,f} =
-\semTrace\,\semTm f$.
+i.e. partial isomorphisms. Here we only display the interpretation of a
+selection of programs, we refer the interested reader to our Agda formalization for a
+complete definition of the interpretation of terms.
+\[
+\begin{array}{ll}
+\multicolumn{2}{l}{\semTm - : (A \lr B) \to \semTy A \, \rho \simeq
+  \semTy B \, \rho}  \\
+\semTm {f \oplus g} & = \semTm f +_{\D\simeq} \semTm g \\
+\semTm {f \otimes g} & = \semTm f \times_{\D\simeq} \semTm g \\
+\semTm {\trace\,f} & = \semTrace\,\semTm f
+\end{array}
+\]
 
@@
 $\Inv\Dapprox$, i.e. proofs of weak bisimilarity between two
-morpshisms. Formally, we define an operation:
+morphisms. Formally, we define an operation:
 \[
@@
 
-\section{Conclusions}
+\section{Conclusions}\label{sec:conclusions}
 In this paper, we have extended the work of Carette and Sabry~\cite{CaretteS16}
@@
 Section \ref{sec:elgot} for the delay monad, which is a simple corecursive
-definition. The Elgot monad structure of $\mathsf{PMC}$ follows from its
+definition. The complete Elgot monad structure of $\mathsf{PMC}$ follows from its
 Kleisli category being a join restriction category, so iteration is defined in

### 2019-06-28 mpc19/paper.tex
Some corrections as suggested by the reviewers.

@@
 isomorphisms and combinators corresponding to those found in dagger-traced
-$\omega$-continuous rig categories~\cite{Karvonen19}. Though superficially
-simple, \Pio{} is expressive enough as a metalanguage to give semantics to the
-typed reversible functional programming language Theseus~\cite{JamesS14}.
-
-In \cite{CaretteS16}, the authors studied the equivalences of isomorphisms in
-the finitary fragment of \Pio{} (i.e., without recursive types and iteration
-via the trace combinator), and showed that these equivalences could be
-adequately described by another combinator calculus of equivalences of
+$\omega$-continuous rig categories~\cite{Karvonen19} (a kind of dagger category
+with a trace, monoidal sums $\oplus$ and products $\oplus$ such that they form
+a rig structure, and fixed points of the functors formed from the rig
+structure). Though superficially simple, \Pio{} is expressive enough as a
+metalanguage to give semantics to the typed reversible functional programming
+language Theseus~\cite{JamesS14}.
+
+In \cite{CaretteS16}, Carette and Sabry studied the equivalences of 
+isomorphisms in the finitary fragment of \Pio{} (i.e., without recursive types
+and iteration via the trace combinator), and showed that these equivalences
+could be adequately described by another combinator calculus of equivalences of
 isomorphisms, in sum yielding a two-level calculus of isomorphisms and
 equivalences of isomorphisms. In this paper, we build on this work to produce a
-(fully formalized) two-level calculus for full \Pio{} (supporting both recursive
-types and iteration) via the delay monad, using insights gained from the
-study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well
-as of join inverse categories in which reversible iteration may be
+(fully formalized) two-level calculus for full \Pio{} (supporting both
+recursive types and iteration) via the delay monad, using insights gained from
+the study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as
+well as of join inverse categories in which reversible iteration may be
 modelled~\cite{KaarsgaardAG17}.
@@
 It is possible to show that every term $f$ has $\dagger\,f$ as its partial
-inverse, and since the final equivalence of Figure~\ref{fig:programequivs} is
-equivalent to stating that partial inverses are unique~\cite{Kastl79}, it
-follows that the types and terms of \Pio\ form an inverse category. The proof
+inverse. The notion of partial inverse used here comes from the study of
+\emph{inverse categories} (see \cite{Kastl79}) and amounts to saying that
+$\dagger\,f$ is the \emph{unique} map that undoes everything which $f$ does
+(unicity of partial inverses follows by the final equivalence of
+Figure~\ref{fig:programequivs}, see~\cite{Kastl79}). Note that this is
+different from requiring that $f$ is an isomorphism in the usual sense, as
+$\dagger\,f \bullet f$ is not going to be the identity when $f$ is only
+partially defined, though it will behave as the identity on all points where
+$f$ is defined.
+
+The proof that every term has $\dagger\,f$ as its partial inverse proof
 proceeds by induction on $f$.
@@
 issue with these concrete formulations of partiality, but rather with the
-interaction of partiality (in the sense of restriction categories) and limits
+interaction of partiality (in the sense of restriction categories, a kind of
+categories of partial maps) and limits
 in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is
@@
 \[
-\semTrace : A + C\simeq B + C \to A \simeq C
+\semTrace : A + C\simeq B + C \to A \simeq B
 \]

### 2019-07-01 mpc19/paper.tex
halfway to final version

@@
 $\omega$-continuous rig categories~\cite{Karvonen19} (a kind of dagger category
-with a trace, monoidal sums $\oplus$ and products $\oplus$ such that they form
+with a trace, monoidal sums $\oplus$ and products $\otimes$ such that they form
 a rig structure, and fixed points of the functors formed from the rig
@@
 Throughout the paper, we reason constructively in Martin-L\"of type
-theory. Classically, the delay monad is isomorphic to the maybe monad
+theory. Classically, the delay monad (quotiented by
+weak bisimilarity)
+is isomorphic to the maybe monad
 $\Maybe\,X \equiv X + 1$,
@@
 not terminate as well. If $f\,a$ terminates, there are two
-possibilities: either $f\,a$ converges to $\inr\,b$, in which case
+possibilities: either $f\,a$ converges to $\inl\,b$, in which case
 $\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
-to $\inl\,a'$, in which case we repeat the procedure by replacing $a$
+to $\inr\,a'$, in which case we repeat the procedure by replacing $a$
 with $a'$. Notice that in the latter case we also add one occurrence
@@
 mutually inductively defined together with the operation $\semMu -$
-giving semantics to the $\mu$ type former.
+giving semantics to the $\mu$ type former. Remember that, given $A :
+\Ty\,(n+1)$ and $B : \Ty\,n$, we write $\sub\,A \,B$ for the
+substituted type $A[B/X]$, where $X$ corresponds to the $(n+1)$-th
+variable in context.
+
 \[
@@
 \qquad \qquad
-\infer{\mathsf{semFold} \,x : \semMu A\,\rho}{x : \semTy A \, \rho}
+\infer{\mathsf{semFold} \,x : \semMu {A}\,\rho}{x : \semTy
+  {\sub\,A\,(\mu A)} \, \rho}
 \]

### 2019-07-01 mpc19/paper.tex
extra paragraph on intuition for trace for delay monad

@@
 \cite{GoncharovSRJ18,GoncharovMR16}.
+
+The uniform iteration operator introduces a notion of feedback,
+typically used to model control flow operations such as while
+loops. In the Kleisli category of the delay monad, this operation can
+be intuitively described as follows: We can apply a function
+$f : A \to B + A$ on an input $a$ and either produce an element
+$b : B$, or produce a new element $a' : A$ which can be fed back to
+$f$. This operation can be iterated, and it either terminates
+returning a value in $B$ or it goes on forever without producing any
+output. This form of iteration is ``unguarded'' because it allows the
+possibility of divergence. The trace operator can then be seen as a
+particular form of iteration where, given a function
+$f : A + C \to B + C$, which can be decomposed as
+$\mapL f : A \to B + C$ and $\mapR f : C \to B + C$, we first apply
+$\mapL f$ on an input $a : A$, and, if the latter operation produces a
+value $c : C$, we continue by iterating $\mapR f$ on $c$. Notice that
+the notion of trace can be generally defined in monoidal categories
+where the monoidal structure is not necessarily given by coproducts,
+and it has been used to model other things besides iteration, such as
+partial traces in vector spaces \cite{JoyalSV96}, though this use
+falls outside of the scope of this paper.
 

### 2019-07-01 mpc19/paper.tex
Minor change.

@@
 be intuitively described as follows: We can apply a function
-$f : A \to B + A$ on an input $a$ and either produce an element
+$f : A \to B + A$ on an input $a : A$ and either produce an element
 $b : B$, or produce a new element $a' : A$ which can be fed back to

### 2019-07-13 mpc19/paper.tex
small fixes

@@
 is isomorphic to the maybe monad
-$\Maybe\,X \equiv X + 1$,
+$\Maybe\,X = X + 1$,
 and thus just a complication of something that can be expressed much
@@
 equal. Agda natively supports UIP, so we have to manually switch it
-using the \verb|without-K| option.
+off using the \verb|without-K| option.
 
@@
 symmetric monoidal structure, with terms $\lambda_\oplus$,
-$\alpha_\oplus$ and $\sigma_\otimes$ as structural
+$\alpha_\oplus$ and $\sigma_\oplus$ as structural
 morphisms. Similarly for $(\I,\otimes)$. Moreover $\otimes$
@@
 
-The proof that every term has $\dagger\,f$ as its partial inverse proof
+The proof that every term has $\dagger\,f$ as its partial inverse
 proceeds by induction on $f$.
@@
 $f \, a \approx g \, a$, for all inputs $a$. When this is the case, we
-also write $f \approx g$.
+also write $f \approx g$. The operation $\bind$ is compatible with
+weak bisimilarity, in the sense that $\bind\,f_1\,x_1 \approx
+\bind\,f_2\,x_2$ whenever $f_1 \approx f_2$ and $x_1 \approx x_2$.
 
@@
 explanation. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
-not terminate as well. If $f\,a$ terminates, there are two
+not terminate either. If $f\,a$ terminates, there are two
 possibilities: either $f\,a$ converges to $\inl\,b$, in which case
@@
 comparatively less is known about this monad on other categories. It could be
-interesting to study under which conditions its iterator exist -- e.g., whether
+interesting to study under which conditions its iterator exists -- e.g., whether
 this is still the case when $\Set$ is replaced with an arbitrary topos. Another
